---
title: "ECON 712B - Problem Set 2"
author: "Alex von Hafften^[I worked on this problem set with a study group of Michael Nattinger, Andrew Smith, and Ryan Mather. I also discussed problems with Emily Case, Sarah Bass, and Danny Edgel.]"
date: "11/19/2020"
output: pdf_document
header-includes:
- \newcommand{\N}{\mathbb{N}}
- \newcommand{\Z}{\mathbb{Z}}
- \newcommand{\R}{\mathbb{R}}
- \newcommand{\Q}{\mathbb{Q}}
- \newcommand{\Lf}{\mathcal{L}}
- \newcommand{\graph}{\text{graph}}
- \usepackage{bm}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

1. In class, we only considered the growth model with inelastic labor supply. This problem relaxes that restriction. Consider the benchmark neoclassical growth model, with production function: $Y_t = F(K_t, A_t N_t)$ where $Y_t$ is output, $K_t$ is capital, $A_t$ is technology, and $N_t$ is labor, and $F$ has constant returns to scale and satisfies the usual assumptions. Technology grows exogenously at rate $g$: $A_{t+1} = (1+g) A_t$.  Capital depreciates at rate $\delta$ so (imposing the aggregate feasibility condition) we can write the law of motion for the capital stock as: $K_{t+1} = (1 - \delta) K_t + Y_t - C_t$. The representative household has time additive preferences given by: $\sum_{t=0}^\infty \beta^t u(C_t, 1 - N_t)$.  The population size is fixed, but the labor input $N_t \in [0,1]$ is now endogenous. This problem will consider the existence of a balanced growth path, which is defined as an equilibrium allocation where consumption, capital, wages $W_t$, and output all grow at the same constant rate, while interest rates $r_t$ and labor $N_t$ are constant.

(a) From conditions characterizing the equilibrium, find a system of equations that the endogenous variables $C_0$, $N_0$, $W_0$, $r_0$ must solve in a balanced growth path. (Initial capital $K_0$ is given.)

In a competitive equilibrium, households optimize, firms optimize, and markets clear.\footnote{Note on notation: The original problem set specifies $w_t$ as the wages that grow over time.  Here, $W_t$ is the real wage rate on labor $N_t$ and $w_t$ is the real wage rate on effective labor $A_tN_t$. In addition, $r_t$ is the real rental rate on capital.}

The household problem is 

\begin{align*}
\max_{\{(C_t, N_t, K_t, I_t)\}_{t=0}^\infty} & \sum_{t=0}^\infty \beta^t u(C_t, 1 - N_t) \\
\text{s.t. } & \sum_{t=0}^\infty p_t (C_t + I_t) = \sum_{t=0}^\infty p_t (r_t K_t + W_t N_t) + \pi_t \\
& K_{t+1} = (1 - \delta) K_t + I_t
\end{align*}

The firm problem is

$$
\max_{\{K_t^d, N_t^d\}_{t=0}^\infty} \sum_{t=0}^\infty p_t(F(K_t^d, A_t N_t^d) - r_tK_t^d - W_tN_t^d)
$$

Transform consumption, capital, investment, and wages by dividing by technology: $c_t = \frac{C_t}{A_t}$, $k_t = \frac{K_t}{A_t}$, $i_t = \frac{I_t}{A_t}$, and $w_t = \frac{W_t}{A_t}$.

\pagebreak

The firm problem becomes:

$$
\max_{\{k_t^d, N_t^d\}_{t=0}^\infty} \sum_{t=0}^\infty p_t(F(A_tk_t^d, A_t N_t^d) - r_t A_t k_t^d - A_t w_t N_t^d)
$$

FOC [$k_t^d$]:\footnote{Note on notation: $F_1$ is the derivative with respect to the first argument of $F$ and $F_2$ is the derivative with respect to the second argument of $F$.}

$$
p_tF_1(A_tk_t^d, A_t N_t^d)A_t =  p_tA_tr_t \implies F_1(A_tk_t^d, A_t N_t^d) = r_t 
$$ 

FOC [$N_t^d$]:

$$
p_tF_2(A_tk_t^d, A_t N_t^d)A_t = p_tA_tw_t \implies F_2(A_tk_t^d, A_t N_t^d) = w_t
$$

Since $F$ has CRS $\implies F(K, N) = KF_1(K, N) + NF_2(K, N)$ by Euler's Theorem. Thus, for all $t$,

\begin{align*}
\pi_t 
&= p_t(F(A_tk_t^d, A_t N_t^d) - r_t A_t k_t^d - A_t w_t N_t^d) \\
&= p_t(A_tk_t^dF_1(A_tk_t^d, A_t N_t^d) + A_t N_t^dF_2(A_tk_t^d, A_t N_t^d) - r_t A_t k_t^d - A_t w_t N_t^d) \\
&= p_t(A_t k_t^d r_t + A_t N_t^dw_t - r_t A_t k_t^d - A_t w_t N_t^d) \\
&= 0
\end{align*}

The household problem becomes:

\begin{align*}
\max_{\{(c_t, N_t, i_t, k_t)\}_{t=0}^\infty} & \sum_{t=0}^\infty \beta^t u(A_tc_t, 1-N_t) \\
\text{s.t. } & \sum_{t=0}^\infty p_t (A_t c_t + A_t i_t) = \sum_{t=0}^\infty p_t (r_t A_t k_t + w_t A_t N_t) + (0) \\
& A_{t+1}k_{t+1} = (1 - \delta) A_{t}k_{t} + A_{t}i_{t} \\
\implies
\max_{\{(c_t, N_t, i_t, k_t)\}_{t=0}^\infty} & \sum_{t=0}^\infty \beta^t u(A_tc_t, 1-N_t) \\
\text{s.t. } & \sum_{t=0}^\infty p_t (c_t + (1+g)k_{t+1} - (1 - \delta) k_{t} - r_t k_t - w_t N_t) = 0
\end{align*}

Define Legrangian:

$$
\Lf 
= \sum_{t=0}^\infty \beta^t u(A_tc_t, 1-N_t) 
+ \lambda\Bigg[\sum_{t=0}^\infty p_t (c_t + (1+g)k_{t+1} - (1 - \delta) k_{t} - r_t k_t - w_t N_t)\Bigg]
$$

\pagebreak

FOC [$c_t$]:\footnote{Note on notation: $u_1$ is the derivative with respect to the first argument of $u$ and $u_2$ is the derivative with respect to the second argument of $u$.}

\begin{align*}
0 &= \beta^t u_1(A_tc_t, 1-N_t)A_t + \lambda p_t \\
\implies
-\lambda &= \frac{\beta^t u_1(A_tc_t, 1-N_t)A_t}{p_t}
\end{align*}

FOC [$N_t$]:

\begin{align*}
0 &= \beta^t u_2(A_tc_t, 1-N_t) (-1) - \lambda p_tw_t \\
\implies
-\lambda &=  -\frac{\beta^t u_2(A_tc_t, 1-N_t)}{p_tw_t}
\end{align*}

Combining the FOCs with respect to $c_t$ and $N_t$, we get the following relationship:

\begin{align*}
\frac{\beta^t u_1(A_tc_t, 1-N_t)A_t}{p_t} &= -\frac{\beta^t u_2(A_tc_t, 1-N_t)}{p_tw_t} \\
\implies
u_1(A_tc_t, 1-N_t) &= -\frac{1}{A_t w_t}u_2(A_t c_t, 1-N_t) \\
\end{align*}

In equilibrium, markets clear:

\begin{align*}
k_t &= k_t^d &(\text{Capital Market})\\
N_t &= N_t^d &(\text{Labor Market})\\
F(A_tk_t, A_tN_t) &= (c_t + k_t)A_t &(\text{Goods Market})
\end{align*}

Finally, the law of capital can be rewritten with our transformed variables:

\begin{align*}
k_{t+1}A_{t+1} &= (1 - \delta) k_tA_t + F(A_tk_t, A_tN_t) - c_tA_t \\
\implies k_{t+1}(1+g) &= (1 - \delta) k_t + F(k_t, N_t) - c_t
\end{align*}

On a balanced growth path, we know that $k_{t+1} = k_t$:

$$
k_t (g + \delta) + c_t=  F(k_t, N_t) 
$$

At $t=0$, there are four equations that characterize the four unknown variables $(C_0, N_0, W_0, r_0)$:

\begin{align}
F_1(K_0, A_0 N_0) &= r_0\\
F_2(K_0, A_0 N_0) &= \frac{W_0}{A_0}\\
u_1(C_0, 1 - N_t) &= -\frac{1}{W_0} u_2(C_0, 1-N_0)\\
K_0 (g + \delta) + C_0 &= A_0F(K_0, N_0)
\end{align}

(b) Show that if preferences are of the form: $u(C, 1-N) = \begin{cases} \frac{C^{1-\gamma}}{1-\gamma}h(1-N), \gamma > 0, \gamma \neq 1 \\ \log C + h(1-N), \gamma = 1 \end{cases}$ for some function $h$, then there will be a balanced growth path.

If $\gamma \neq 1$, power utility preferences imply:

\begin{align*}
u_1(C, 1-N) &= C^{-\gamma}h(1-N) \\
u_2(C, 1-N) &= \frac{C^{1-\gamma}}{1-\gamma}h'(1-N)
\end{align*}

The FOCs with respect to $c_t$ and $c_{t+1}$ of the Legrangian to solve the household problem imply a consumption Euler equation:

\begin{align*}
\frac{\beta^t u_1(A_tc_t, 1-N_t)A_t}{p_t} &= \frac{\beta^{t+1} u_1(A_{t+1}c_{t+1}, 1-N_{t+1})A_{t+1}}{p_{t+1}} \\ 
\implies 
u_1(A_tc_t, 1-N_t) &= \frac{\beta (1 + g )}{q_{t+1}}u_1(A_{t+1}c_{t+1}, 1-N_{t+1})
\end{align*}

where $q_{t+1} = \frac{p_{t+1}}{p_t}$.  The functional form for utility implies:

\begin{align*}
(A_t c_t)^{-\gamma} h(1-N_t) &= \frac{\beta (1 + g )}{q_{t+1}}(A_{t+1}c_{t+1})^{-\gamma}h(1-N_{t+1}) \\
c_t^{-\gamma} h(1-N_t) &= \frac{\beta (1 + g )}{q_{t+1}}(1+g)^{-\gamma}c_{t+1}^{-\gamma}h(1-N_{t+1})
\end{align*}

On the balanced growth, $\bar{N} = N_t = N_{t+1}$, $\bar{c} = c_t = c_{t+1}$, and $\bar{q} = q_{t+1}$:

$$
\bar{q}= \beta (1 + g )^{1-\gamma}
$$

The FOCs with respect to $N_t$ and $N_{t+1}$ of the Legrangian to solve the household problem imply a labor supply Euler equation:

\begin{align*}
\frac{\beta^t u_2(A_tc_t, 1-N_t)}{p_tw_t} &=  \frac{\beta^{t+1} u_2(A_{t+1}c_{t+1}, 1-N_{t+1})}{p_{t+1}w_{t+1}} \\ 
\implies 
u_2(A_tc_t, 1-N_t) &= \frac{\beta}{q_{t+1}}\frac{w_{t}}{w_{t+1}} u_2(A_{t+1}c_{t+1}, 1-N_{t+1})
\end{align*}

The functional form for utility implies:

\begin{align*}
\frac{(A_tc_t)^{1-\gamma}}{1-\gamma}h'(1-N_t) &= \frac{\beta}{q_{t+1}}\frac{w_{t}}{w_{t+1}} \frac{(A_{t+1}c_{t+1})^{1-\gamma}}{1-\gamma}h'(1-N_{t+1}) \\
c_t^{1-\gamma}h'(1-N_t) &= \frac{\beta}{q_{t+1}}\frac{w_{t}}{w_{t+1}} c_{t+1}^{1-\gamma}(1+g)^{1-\gamma}h'(1-N_{t+1}) \\
\end{align*}

On the balanced growth, $\bar{N} = N_t = N_{t+1}$, $\bar{c} = c_t = c_{t+1}$, and $\bar{q} = q_{t+1}$:

$$
\bar{q} \frac{w_{t}}{w_{t+1}} = \beta (1+g)^{1-\gamma} \\
\implies w_{t} = w_{t+1}
$$

The Legrangian also implies a "no arbitrage condition", by taking the derivative with respect to $k_{t+1}$:

\begin{align*}
0 &= \lambda p_t (1+g) - \lambda p_{t+1}(1 - \delta) - \lambda p_{t+1} r_{t+1} \\
\implies 
q_{t+1} &= \frac{1+g}{(1 - \delta) + r_{t+1} }
\end{align*}

On the balanced growth, $\bar{q} = q_{t+1}$ and $\bar{r} = r_{t+1}$:

\begin{align*}
\beta (1 + g )^{1-\gamma} = \bar{q} &= \frac{1+g}{(1 - \delta) + \bar{r} } \\
\implies 
\bar{r} &= \frac{(1 + g)^{\gamma}}{\beta} - (1 - \delta) 
\end{align*}

The firm problem implies the values of $\bar{k}$ and $\bar{N}$:

\begin{align*}
F_1(\bar{k}, \bar{N}) &= \bar{r} \\
F_2(\bar{k}, \bar{N}) &= \bar{w} \\
\end{align*}

Finally, the law of motion for capital implies the value of $\bar{c}$:

$$
\bar{c} =  F(\bar{k}, \bar{N}) - \bar{k} (g + \delta)
$$

Therefore, a balanced growth path exists.

(c) Can we characterize the qualitative dynamics using a phase diagram in the same way that we did in the case of inelastic labor supply? For example, suppose $u(C, 1 - N) = \log C + h(1 - N)$, that we are on a balanced growth path and then there is an increase in the rate of depreciation $\delta$. Can you say what happens both upon impact of the shock and in the long run?

No, since labor, consumption and capital all endongenous.  All three can change in response to an increase in $\delta$. We need to know the production function to determine the impact of the shock in the short and long run.

(d) Now suppose that $h$ is a constant function, so that labor is inelastically supplied, and suppose $\gamma > 1$. Show that we can summarize the equilibrium as a system of equations governing the evolution of consumption and capital per unit of effective labor: $c_t = C_t/A_t$ and $k_t = K_t/A_t$. Find the balanced growth path levels of $c_t$ and $k_t$.

If $h$ is a constant function, then $N_t = 1$ for all $t$.  The equations from part (b) hold here, in particular:

\begin{align*}
c_t^{-\gamma} &= \frac{\beta (1 + g )}{q_{t+1}}(1+g)^{-\gamma}c_{t+1}^{-\gamma} \\
q_{t+1} &= \frac{1+g}{(1 - \delta) + r_{t+1} } \\
k_{t+1}(1 + g) &= (1 - \delta) k_t + F(k_t, N_t) - c_t
\end{align*}



(e) Now suppose the economy is on the balanced growth path, and then there is a fall in the rate of technological change $g$. By analyzing the qualitative dynamics of the economy, discuss what happens to $c_t$ and $k_t$ at the time of the change and in the long run.

(f) For a marginal change in $g$, find an expression showing how the fraction of output saved on the balanced growth path changes. Does savings increase or decrease? Consider first a general production function, and then specialize to Cobb-Douglas production: $F(K, N) = K^\alpha N^{1-\alpha}$.

\pagebreak

2. At any date $t$, a consumer has $x_t$ units of a non-storable good. He can consume $c_t \in [0, x_t]$ of this stock, and plant the remaining $x_t - c_t$ units. He wants to maximize: $E \sum_{t=1}^\infty \beta^t \frac{c_t^{1-\gamma}}{1-\gamma}$ where $0 < \gamma < 1$ and $0 < \beta < 1$. Goods planted at date $t$ yield $A_t(x_t - c_t)$ as of the beginning of period $t + 1$, where $A_t$ is a sequence of i.i.d. random variables that take the values of $0 < A_h < 1/\beta$ with probability $\pi$ and $A_l \in (0, A_h)$ with probability $1 - \pi$.

(a) Formulate the consumer’s utility maximization problem in the space of shock contingent consumption sequences. Exactly what is this space? Exactly what does the expectations operator $E(\cdot)$ mean here? Be explicit.

Thus, the consumer's expected utility maximization problem is

$$
\max_{\{c_t\}_{t=0}^\infty} E_0 \Bigg[ \sum_{t=0}^\infty \frac{\beta^t c_t^{1-\gamma}}{1- \gamma} \Bigg] \\
\text{s.t. } x_t \le A_{t-1}(x_{t-1} - c_{t-1})
$$

The probability space is the sample space $\Omega = \{A_h, A_l\}$, the $\sigma$-algebra $\mathcal{F} = \{\emptyset, \{A_h\}, \{A_l\}, \{A_h, A_l\}\}$, and the probability measure $Q: \mathcal{F} \to [0, 1]$ such that $Q(A_h) = \pi$, $Q(A_l) = 1-\pi$, and $Q(\emptyset) = Q(\{A_h, A_l\}) = 0$. Equivalently, as discussed in lecture, we can represent $Q$ as a Markov transition matrix that does not depend on the current state: 

$$
\begin{bmatrix} \pi & 1- \pi\\ \pi & 1- \pi \end{bmatrix}
$$

The expectations operator averages total discounted utility of different shock contingent consumption sequences weighted by the probability of those shocks.

(b)  State the Bellman equation for this problem. It is easiest to have the consumer choose savings $s_t = x_t - c_t$. Argue that the relevant state variable for the problem is the cum-return wealth $A_{t-1}s_{t-1}$. Prove that the optimal value function is continuous, increasing, and concave in this state. How can you handle the unboundedness of the utility function?

Define $s_t = x_t - c_t = A_{t-1}s_{t-1} - c_t$. Furthermore, the probability measure does not depend on the current state for future shock. Thus, the consumption-savings decision by the household in $t$ depends only on $A_{t-1}s_{t-1}$. Therefore, the Bellman equation is:

$$
V(As) = \max_{s'} \Bigg\{ \frac{(As - s')^{1-\gamma}}{1-\gamma} + \beta E\Big[V(A' s')\Big] \Bigg\}
$$

We can substitute in the probability-weighted average of the continuation value:

$$
V(As) = \max_{s'} \Bigg\{ \frac{(As - s')^{1-\gamma}}{1-\gamma} + \beta\Big[\pi V(A_h s') + (1-\pi)V(A_l s')\Big] \Bigg\}
$$

Define the feasible cooresponendence as the values of savings that result in nonnegative consumption: $\Gamma(x) = [0, x]$. Define the objective function $F(x, y) = u(x-y) = \frac{(x - y)^{1-\gamma}}{1-\gamma}$. The optimal value function is unique, continuous, increasing, and concave based on the following conditions:

- ($\Gamma 1$): $\Gamma$ nonempty, compact-valued, and continuous.

- ($\Gamma 2$): $\Gamma$ is monotone (i.e., $x \le x' \implies \Gamma(x) \subseteq \Gamma(x')$).

- ($\Gamma 3$): $\Gamma$ is convex.

- ($F 1$): $F$ and continuous with $0 < \beta < 1$. Although $u$ (and by extension $F$) is not bounded, the value function is guaranteed to be finite because $A_h \beta < 1$ and $u$ is concave. In other words, after factoring in return on savings, the discounted utility of a unit of consumption descreases over time.

- ($F 2$) $\forall y$, $F( \cdot, y)$ is strictly increasing.

- ($F 3$) $F$ is strictly concave in $(x, y)$:

$$
\frac{\partial^2 F}{\partial^2 x} = \frac{\partial^2 F}{\partial^2 y} = -\gamma(x - y)^{-\gamma-1} < 0
$$

- ($F 4$) $F$ is continuously differentiable on the interior of the feasible set. 

(c) Solve the Bellman equation and obtain the corresponding optimal policy function. (Hint: guess that the optimal function consists of saving a constant fraction of wealth.)

FOC [$s'$]:

$$
(As - s')^{-\gamma} =  \beta\pi V(A_h s')A_h + \beta(1-\pi)V(A_l s')A_l
$$

Envelope Condition:

$$
V'(As) = (As - s')^{-\gamma} 
$$

FOC and envelope condition combine to result in a Euler equation:

$$
(As - s')^{-\gamma} =  \beta\pi (A_hs' - s''_h)^{-\gamma}A_h + \beta(1-\pi)(A_ls' - s''_l)^{-\gamma}A_l
$$

where $s''_h$ is the savings in two periods if $A' = A_h$ and $s''_l$ is the savings in two periods if $A' = A_l$. Guess that the optimal function is saving a constant fraction of wealth: $s' = f(As) = pAs$ where $p \in \R_+ \implies s''_h = p^2AA_hs$ and $s''_l = p^2AA_ls$

\begin{align*}
(As - pAs)^{-\gamma} &= \beta\pi (A_hpAs - p^2AA_hs)^{-\gamma}A_h + \beta(1-\pi)(A_lpAs - p^2AA_ls)^{-\gamma}A_l \\
(As)^{-\gamma}(1 - p)^{-\gamma} &= \beta\pi (A_hpAs)^{-\gamma}(1-p)^{-\gamma}A_h + \beta(1-\pi)(1-p)^{-\gamma}(A_lpAs)^{-\gamma}A_l \\
1 &= \beta\pi (A_hp)^{-\gamma}A_h + \beta(1-\pi)(A_lp)^{-\gamma}A_l \\
p^{\gamma} &= \beta\pi A_h^{1-\gamma} + \beta(1-\pi)A_l^{1-\gamma} \\
p &= (\beta\pi A_h^{1-\gamma} + \beta(1-\pi)A_l^{1-\gamma})^{(1/\gamma)}
\end{align*}

(d)  How do you know that the consumption sequence generated by this policy function is the unique solution of the original sequence problem?

Based on the assumptions outlined in part (b), this policy function solve the original sequence problem and it is unique.

\pagebreak

3. This problem considers the computation of the optimal growth model. An infinitely lived representative household owns a stock of capital which it rents to firms. The household’s capital stock $K$ depreciates at rate $\delta$. Households do not value leisure and are endowed with one unit of time each period with which they can supply labor $N$ to firms. They have standard time additive expected utility preferences with discount factor $\beta$ and period utility $u(c)$. Firms produce output according to the production function $zF(K, N)$ where $z$ is the level of technology.

(a) First, write a computer program that solves the planners problem to determine the optimal allocation in the model. Set $\beta = 0.95$, $\delta = 0.1$, $z = 1$, $u(c) = c^{1-\gamma}/(1 - \gamma)$ with $\gamma = 2$, and $F(K, N) = K^{0.35}N^{0.65}$. Plot the optimal policy function for $K$ and the phase diagram with the $\Delta K = 0$ and $\Delta c = 0$ lines along with the saddle path (which is the decision rule $c(K)$).

The planners problem is very similar to the one discussed in lecture 4:

\begin{align*}
\max_{\{c_t, n_t, k_{t+1}\}_{t=0}^\infty} & \sum_{t=0}^\infty \beta^t u(c_t) \\
\text{s.t. } & c_t + k_{t+1} - (1-\delta)k_t \le zF(k_t, n_t) \\
& 0 \le n_t \le 1 \\
& c_t \ge 0 
\end{align*}

It is optimal for households to supply $n_t = 1$ for all $t$ because leisure is not valued.  Define $F(k_t, 1) = f(k_t)$.  The recursive formulation of the planners problem is

$$
V(k) = \max_{k'} \{u(zf(k) + (1-\delta)k - k') + \beta V(k')\}
$$

FOC [$k'$]:

$$
0=u'(zf(k) + (1-\delta) k - k')(-1) + \beta V'(k')
$$

Envelope condition:

$$
V'(k) = u'(zf(k) + (1-\delta) k - k')(zf'(k)+(1-\delta))
$$

Imply an Euler condition:

$$
u'(zf(k) + (1-\delta) k - k') = \beta u'(zf(k') + (1-\delta) k' - k'')(zf'(k')+(1-\delta)) 
$$

The Euler equation and the law of motion of capital are two difference equation for the two unknowns:

\begin{align}
k' &= zf(k) + (1-\delta)k - c \\
u'(c) &= \beta u'(c')(zf'(k')+(1-\delta))
\end{align}

In a steady state $\bar{k} = k = k'$ and $\bar{c} = c = c'$:

\begin{align*}
u'(\bar{c}) &= \beta u'(\bar{c})(zf(\bar{k})+(1-\delta)) \\
\implies f'(\bar{k}) &= \frac{\beta^{-1} - (1-\delta)}{z} \\
\bar{c} &= zf(\bar{k}) -\delta \bar{k} 
\end{align*}

Plugging in the provided functional forms:

\begin{align*}
\bar{k} &= \Bigg(\frac{\beta^{-1} - (1-\delta)}{0.35z}\Bigg)^{(-1/0.65)} \\
\bar{c} &= z\bar{k}^{0.35} - \delta \bar{k} 
\end{align*}

For $\Delta k = 0$, equation (5) implies:

$$
k = zf(k) + (1-\delta)k - c \implies c = zf(k) - \delta k
$$

Plugging in the provided functional forms:

$$
c = zk^{0.35} - \delta k
$$

For $\Delta c = 0$, equations (5) and (6) imply:

$$
f'(zf(k) + (1-\delta)k - c) = \frac{\frac{1}{\beta}-(1-\delta)}{z}
$$

Plugging in the provided functional forms:

\begin{align*}
0.35(zk^{0.35} + (1-\delta)k - c)^{-0.65} &= \frac{\frac{1}{\beta}-(1-\delta)}{z} \\
\implies  c &= zk^{0.35} + (1-\delta)k - \Bigg(\frac{\frac{1}{\beta}-(1-\delta)}{0.35z}\Bigg)^{1/(-0.65)}
\end{align*}

```{r setup_model, echo = FALSE}
# parameters
beta <- 0.95
delta <- 0.1
z <- 1
gamma <- 2
cd_k <- 0.35 # cobb-douglas coefficient on capital
cd_l <- 0.65 # cobb-douglas coefficient on labor

# capital grid
min_k <- 0.005
max_k <- 5
inc_k <- 0.005
k <- seq(min_k, max_k, by = inc_k)
n <- length(k)
```

```{r solve_ss, echo = FALSE}
solve_ss <- function() {
  # Steady state values
  k_bar <<- ((beta^(-1)-(1-delta))/(cd_k * z))^(1/(cd_k-1))
  c_bar <<- z * k_bar^cd_k - delta * k_bar

  # phase diagram
  k_no_change <<- z * k ^ cd_k - delta * k
  c_no_change <<- z * k ^ cd_k + (1-delta) * k - k_bar
}
```

```{r solve_bellman_equation, echo = FALSE}
# Solves bellman equation based on parameters above.
solve_bellman <- function() {
  ones <- rep(1, times=n)
  
  # initialize value grid and decision rule grid
  v <- rep(0, times=n)
  decision <- rep(0, times=n)
  
  # create consumption and utility matrixes
  # (columns are different values of k; rows are values of k')
  c_matrix <- z*(ones %*% t(k))^cd_k + (1-delta)*(ones %*% t(k)) - k %*% t(ones)
  utility_matrix <- (c_matrix^(1-gamma))/(1-gamma)
  utility_matrix[c_matrix < 0|!is.finite(utility_matrix)] <- -1000
  
  test <- 10
  while (test != 0) {
    # create value matrix for all different k and k' values
    value_matrix <- utility_matrix + beta * v %*% t(ones)
    
    # find column max with max utility
    tv <- apply(value_matrix, 2, max)
    tdecision <- apply(value_matrix, 2, which.max)
    
    # loop ends if no changes from previous iteration
    test <- max(tdecision - decision)
    
    # update value grid and decision rule
    decision <- tdecision
    v <- tv
  }
  
  # capital and consumption decision rule
  policy_function <<- tibble(k = k, k_prime = min_k + decision*inc_k) %>%
    mutate(c = z*k_prime^cd_k + (1-delta)*k - k_prime)
  
  v <<- v
}
```

```{r plots_a, echo = FALSE}
solve_ss()

solve_bellman()

plot(x=policy_function$k, y=policy_function$k_prime, type = "l", xaxs = "i", yaxs = "i", ylim=c(0, 5), xlim=c(0, 5), xlab = "k", ylab = "k'", main = "Policy Function: k'(k)")

plot(x=k, y=v, type = "l", xaxs = "i", yaxs = "i", xlim=c(0, 5), main = "Value Function")

plot(1, type = "n", xlim = c(1, 5), ylim = c(0, 3), xlab = "k", ylab = "c", xaxs = "i", yaxs = "i", main = "Phase Diagram")

abline(h=0)

lines(x=k, y=k_no_change, col = "red")
lines(x=k, y=c_no_change, col = "blue")
lines(x=policy_function$k, y=policy_function$c)

points(x=k_bar, y=c_bar, pch = 16)

text(x= 2, y=1.5, expression(paste(Delta, "k = 0")), col = "red")
text(x= 4, y=2, expression(paste(Delta, "c = 0")), col = "blue")
text(x= 1.5, y=.5, "Saddle Path")

```

\pagebreak

(b) Re-do your calculations with $\gamma = 1.01$. What happens to the steady state? What happens to the saddle path? Interpret your answer.

$\gamma$ does not appear in the steady state equations nor the equations consistent with $\Delta k = 0$ and $\Delta c = 0$, so the steady state is unaffected. The saddle path rotates about the steady state with the saddle path higher on the left of the steady state and lower on the right.  With $\gamma$ shifted down from 2 to 1.01, the utility function drops off more slowly as $c$ approaches zero.  Thus, when the economy is below the steady state, the household consumes less with a lower $\gamma$.

```{r plots_b, echo = FALSE, fig.height=4.25}
z <- 1
gamma <- 2

solve_ss()

solve_bellman()

z <- 1
gamma <- 1.01

policy_function_old <- policy_function

solve_bellman()

plot(x=policy_function$k, y=policy_function_old$k_prime, type = "l", xlab = "k", ylab = "k'", main = "Policy Function: k'(k)", xaxs = "i", yaxs = "i", ylim = c(0, 5), xlim = c(0, 5))

abline(h=k_bar, v=k_bar, lty =3)

lines(x=policy_function$k, y=policy_function$k_prime, col = "forestgreen", lty=2)
text(x= 1, y=2.5, expression(paste(gamma, " = 1.01")), col = "forestgreen")
text(x= 1.5, y=.75, expression(paste(gamma, " = 2")))

mtext("k_bar", at = k_bar, side = 1)
mtext("k_bar", at = k_bar, side = 2)

plot(1, type = "n", xlim = c(1, 5), ylim = c(0, 3), xlab = "k", ylab = "c", xaxs = "i", yaxs = "i", main = "Phase Diagram")

abline(h=0)

lines(x=k, y=k_no_change, col = "red")
lines(x=k, y=c_no_change, col = "blue")
lines(x=policy_function$k, y=policy_function$c, col = "forestgreen", lty=2)
lines(x=policy_function_old$k, y=policy_function_old$c)

points(x=k_bar, y=c_bar, pch = 16)

text(x= 2, y=1.5, expression(paste(Delta, "k = 0")), col = "red")
text(x= 4, y=2, expression(paste(Delta, "c = 0")), col = "blue")
text(x= 2, y=.5, expression(paste("Saddle Path (", gamma, " = 1.01)")), col = "forestgreen")
text(x= 2.5, y=.75, expression(paste("Saddle Path (", gamma, " = 2)")))

```

\pagebreak

(c) Now with $\gamma = 2$ assume that there is an unexpected permanent increase of 20% in total factor productivity, so now $z = 1.2$. What happens to the steady state levels of consumption and capital? Assuming the economy is initially in the steady state with $z = 1$, what happens to consumption and capital after the increase in $z$?

Both the steady state level of consumption and capital increases.  After the unexpected permanay increase to total factor productivity, consumption jumps up to be on the new saddle path then the economy moves along the saddle path to the new steady state path. 

```{r plots_c, echo = FALSE}
gamma <- 2
z <- 1

solve_ss()
solve_bellman()

k_bar_old <- k_bar
c_bar_old <- c_bar
k_no_change_old <- k_no_change
c_no_change_old <- c_no_change

policy_function_old <- policy_function

z <- 1.2

solve_ss()
solve_bellman()

transition_path_k <- c(k_bar_old, policy_function[policy_function$k >= k_bar_old & policy_function$k <= k_bar, ]$k) 
transition_path_c <- c(c_bar_old, policy_function[policy_function$k >= k_bar_old & policy_function$k <= k_bar, ]$c) 

plot(1, type = "n", xlim = c(1, 5), ylim = c(0, 3), xlab = "k", ylab = "c", xaxs = "i", yaxs = "i", main = "Phase Diagram")

abline(h=0)

lines(x =transition_path_k, y = transition_path_c, lwd =3, col = "yellow")

lines(x=k, y=k_no_change_old, col = "red")
lines(x=k, y=c_no_change_old, col = "blue")
lines(x=k, y=k_no_change, col = "red", lty=5)
lines(x=k, y=c_no_change, col = "blue", lty=5)

lines(x=policy_function_old$k, y=policy_function_old$c)
lines(x=policy_function$k, y=policy_function$c, lty=5)

points(x=k_bar, y=c_bar)
points(x=k_bar_old, y=c_bar_old, pch = 16)

text(x= 2, y=1.5, expression(paste(Delta, "k = 0")), col = "red")
text(x= 4, y=2, expression(paste(Delta, "c = 0")), col = "blue")
text(x= 1.5, y=.5, "Saddle Path")

legend("topleft", c("z = 1", "z = 1.2"), lty = c(1, 5))
```

\pagebreak

## Appendix to Problem 3: R code to solve Bellman Equation

```{r appendix1, ref.label='setup_model', eval = FALSE}
```

```{r appendix2, ref.label='solve_bellman_equation', eval = FALSE}
```
