---
title: "ECON 712B - Problem Set 3"
author: "Alex von Hafften^[I worked on this problem set with a study group of Michael Nattinger, Andrew Smith, and Ryan Mather. I also discussed problems with Emily Case, Sarah Bass, and Danny Edgel.]"
date: "12/3/2020"
output: pdf_document
header-includes:
- \newcommand{\N}{\mathbb{N}}
- \newcommand{\Z}{\mathbb{Z}}
- \newcommand{\R}{\mathbb{R}}
- \newcommand{\Q}{\mathbb{Q}}
- \newcommand{\Lf}{\mathcal{L}}
- \newcommand{\graph}{\text{graph}}
- \usepackage{bm}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

1.  An endowment economy consists of two types of consumers indexed by $i = 1, 2$. There is one nonstorable consumption good. Consumer 1 has initial endowment stream $e^1_0 = 1$ while consumer 2 has initial endowment $e^2_0 = 0$. At some random date $s > 0$ the endowments will switch so $e^1_t = 0$ and $e^2_t = 1$ for all $t \ge s$. Suppose the switch happens with probability $\delta$ each period until it is realized. Both consumers have preferences ordered by $E_0 \sum_{t=0}^\infty \beta^t \log c_t^i$. For this problem you can either consider an Arrow-Debreu market structure or a recursive equilibrium with a sequential market structure. For the Arrow-Debreu market, consider a single market at date zero with claims to consumption in any future date and any state. For the recursive structure, consider a market each period in claims to the two "trees,"" and use backward induction to find the values after and then before the switch of the endowments.

(a) Define a competitive equilibrium for this economy for your choice of market structure, being explicit about the objects which make up the equilibrium and the conditions they must satisfy.

(b) Compute a competitive equilibrium, finding the consumption allocation and the prices (as of date zero) of claims to each consumer’s endowment process.

(c) What is price of a claim to the aggregate endowment? What is the risk-free interest rate?

(d) Now suppose that $\lambda \in (0, 1)$ is the Pareto weight on consumer 1 with $1 - \lambda$ the weight on consumer 2. Formulate the social planner’s problem and characterize the solution of the Pareto optimal allocation for a given $\lambda$.

(e) Relate the competitive equilibrium that you found earlier with the Pareto optima. Is the competitive equilibrium optimal? (Can you find a particular $\lambda$ so that they are equivalent?) For a given $\lambda$, can you find a way to decentralize the Pareto optimum as a competitive equilibrium?

\pagebreak

2. Suppose that a representative agent has preferences: $E \sum_{t=0}^\infty \beta^t u(c_t)$ over the single nonstorable consumption good (“fruit”), where $u(c) = c^{1-\gamma}/(1-\gamma)$ for $\gamma > 0$. Her endowment of the good is governed by a Markov process with transition function $F(x', x)$.

(a) Define a recursive competitive equilibrium with a market in claims to the endowment process ("trees").

Let $a_t$ be the representative agent's claim to "trees" in period $t$, $p_t$ be the price of trees in period $t$, and $s_t$ be the stochastic output from each "trees" in period $t$. The representative agent problem is:

\begin{align*}
\max_{\{c_t, a_{t+1}\}}& E_0 \sum_{t=0}^\infty \beta^t u(c_t) \\
\text{s.t. } c_t + p_ta_{t+1} &\le (p_t + s_t)a_t
\end{align*}

where $s_t \sim F(ds', s)$.  We posit that $p_t = p(s_t)$.  The Bellman equation is

\begin{align*}
V(a, s) &= \max_{\{c, a'\}} \{u(c) + \beta \int V(a', s')F(ds', s)\} \\
\text{s.t. } c + p(s)a' &\le (p(s) + s)a\\
c &\ge 0 \\
0 &\le a' \le 1
\end{align*}

A recursive competitive equilibrium is continuous pricing function $p(s)$ and a continuous and bounded function $V^*(a, s)$ such that (i) $V^*(a, s)$ solves the Bellman equation and (ii) $\forall s$, $V^*(1, s)$ is attained by $c \ge s$ and $s' = 1$.  Rewriting the state variables $\{a, s\}$ as wealth $(p(s)+s)a$ and substituting $c = c(s) = (p(s) + s)a - p(s)a'$, the Bellman equation becomes:

$$
V((p(s)+s)a) = \max_{a'} \{u((p(s) + s)a -p(s)a' ) + \beta \int V((p(s')+s')a')F(ds', s)\}
$$

FOC:

$$
u'(c(s)) p(s) = \beta \int V'((p(s')+s')a')(p(s')+s')F(ds', s)
$$

Envelope condition:

$$
V'((p(s) + s)a) = u'(c(s))
$$

FOC and envelope condition imply an Euler equation:

$$
u'(c(s)) = \beta \int u'(c(s'))\frac{p(s')+s'}{p(s)}F(ds', s)
$$

Defining $R_{t+1} = \frac{p_{t+1} + s_{t+1}}{p_t}$, the Euler equation becomes:

$$
u'(c_t) = \beta E_t [u'(c_{t+1})R_{t+1}]
$$

At equilibrium, $c = s$ and $a' = a = 1$.  Thus the pricing function is determined by:

$$
p(s) = \beta \int \frac{u'(s')}{u'(s)}(p(s')+s')F(ds', s)
$$

With CRRA utility, the pricing function becomes:

$$
p(s) = \beta \int \Bigg(\frac{s}{s'}\Bigg)^{\gamma}(p(s')+s')F(ds', s)
$$

(b) When the consumer has logarithmic utility, $u(c) = \log c$, what is the equilibrium price/dividend ratio of a claim to the entire consumption stream? How does it depend on the distribution of consumption growth?

With log utility, the pricing function becomes:

$$
p(s) = \beta \int \frac{s}{s'}(p(s')+s')F(ds', s)
$$

Thus, the price/dividend ratio is

$$
\frac{p(s)}{s} = \frac{\beta \int \frac{s}{s'}(p(s')+s')F(ds', s)}{s} = \beta \int \frac{s}{s'}\frac{p(s')+s'}{s}F(ds', s) =  \beta \int \frac{s}{s'}R'F(ds', s)
$$

If consumption is growing, the price/dividend is lower than if consumption is shrinking,

$$
c' > c\implies s' > s \implies \downarrow \frac{p(s)}{s}
$$

$$
c' < c \implies s' < s \implies \uparrow \frac{p(s)}{s}
$$

(c) Suppose there is news at time $t$ that future consumption will be higher. How will prices respond to this news? How does this depend on the consumer’s preferences (which are CRRA, but not necessarily log)? Interpret your results.

The pricing function is:

$$
p(s) = E\Bigg[ \Bigg(\frac{c}{c'}\Bigg)^{\gamma}(p(s')+s')\Bigg]
$$

If $c'$ increases, prices drop. A higher $\gamma$ results in a larger drop in price.  For CRRA utility, $1/\gamma$ is the the intertemporal elasticity of substitution (IES).  A higher IES is associated with a stronger preference for consumption smoothing across periods.  If $\gamma$ is lower then $1/\gamma$ is higher, so households have a stronger preference for consumption smoothing.  Thus, the news that future consumption will be higher has less of an effect of today's prices.  If $\gamma$ is higher, then $1/\gamma$ is lower, so households will change their consumption pattern more in response to this news.

\pagebreak

(d) Consider an option which is bought or sold in period $t$ and entitles the current owner to exercise the right to buy one “tree” in period $t + 1$ at the fixed price $\bar{p}$ specified at date $t$. (The buyer may choose not to exercise this option.) Find a formula for the price of this option in terms of the parameters describing preferences and endowments.

The pricing function implies a pricing kernal:

$$
q(s, s') = \beta \frac{u'(s')}{u'(s)}f(s', s)
$$

where $f$ is the PDF that corresponds to $F$, or $F(s', s) = \int_{-\infty}^{s'} f(u, s)du$.  

The call option is in-the-money if the strike price $\bar{p}$ is lower than the future market price $p(s')$. So, the payoff function for the call option is 

$$
g(s') = 
\begin{cases} 
p(s') - \bar{p}, & p(s') > \bar{p}\\
0, & p(s') \le \bar{p}
\end{cases}
$$

Thus, the price of the call option is

\begin{align*}
p^c(s) 
&= \int_{-\infty}^\infty q(s, s') g(s') ds'\\
&= \int_{-\infty}^\infty \beta \frac{u'(s')}{u'(s)} (p(s') - \bar{p}) 1\{p(s') > \bar{p}\} f(s', s)ds'\\
&= \int_{-\infty}^\infty \beta \frac{u'(s')}{u'(s)} (p(s') - \bar{p}) 1\{s' > p^{-1}(\bar{p})\} f(s', s)ds'\\
&= \int_{p^{-1}(\bar{p})}^\infty \beta \frac{u'(s')}{u'(s)} (p(s') - \bar{p}) f(s', s)ds'
\end{align*}

The last equality requires $p(\cdot)$ to strictly monotonically increasing in $s$ and thus invertible. With CRRA utility, the call option price formula becomes:

$$
p^c(s) = \int_{p^{-1}(\bar{p})}^\infty \beta \Bigg(\frac{s}{s'}\Bigg)^{\gamma} (p(s') - \bar{p}) f(s', s)ds'
$$

\pagebreak

3. Suppose consumers have preferences given by: $E \sum_{t=0}^\infty \beta^t \frac{c_t^{1-\gamma}}{1-\gamma}$ with $\beta \in (0,1)$ and $\gamma > 1$. The consumer's budget constraint is: $c_t + a_{t+1} = w l_t + (1 + r) a_t$ where $l_t$ is the labor endowment, $a_t$ is the amount of assets, $c_t \ge 0$ is consumption, $w > 0$ is the (constant) wage, and $r > 0$ is the (constant) real interest rate. The consumer also faces the borrowing constraint $a_t \ge 0$. Suppose that the labor endowment follows a finite-state Markov chain with transition matrix $Q$.

(a) Write down the consumer’s Bellman equation, and state the optimality conditions.

\begin{align*}
V(a, l) &= \max_{(c, a') \in \Gamma(a, l)} \Bigg\{u(c) + \beta \int V(a', l')Q(l, dl')\Bigg\} \\
\implies V(a, l) &= \max \Bigg\{u(w l + (1 + r) a - a') + \beta \int V(a', l')Q(l, dl')\Bigg\}
\end{align*}

FOC:

\begin{align*}
0 &= (-1)u(w l + (1 + r) a - a') + \beta \int V_a(a', l')Q(l, dl')\\
\implies u'(c) &= \beta \int V_a(a', l')Q(l, dl')
\end{align*}

Envelope condition:

\begin{align*}
V_a(a, l) &= u'(w l + (1 + r) a - a')(1+r)\\
\implies V_a(a, l) &= (1+r)u'(c)
\end{align*}

FOC and envelope condition implies a stochastic Euler equation:

\begin{align*}
u'(c) &= \beta (1+r) \int u'(c', l')Q(l, dl') \\
\implies u'(c_t) &= \beta (1+r) E_t u'(c_{t+1})
\end{align*}

Plugging in the functional form for the utility function:

$$
c_t^{-\gamma} = \beta(1+r)E_tc_{t+1}^{-\gamma}
$$

(b) The rest of the question will numerically solve for the policy functions and the stationary distribution. Suppose that $l_t$ takes on the following values: $l_t \in \{0.7, 1.1\}$, with the transition matrix given by: $Q = \begin{bmatrix} 0.85 & 0.15 \\ 0.05 & 0.95 \end{bmatrix}$, where $Q(i, j) = P(l_{t+1} = l_j | l_t = l_i)$. Find the stationary distribution and the unconditional mean of the labor endowment.

The stationary distribution of a Markov Chain is $\pi$ such that $\pi Q = \pi$. Thus we find the eigenvalues and eigenvectors of $Q$.  The characteristic polynomial of $Q$ is 

$$
(0.85 - \lambda)(0.95 - \lambda) - (0.15)(0.05) = \lambda^2 - 1.8 \lambda + 0.8 = (\lambda - 1)(\lambda - 0.8)
$$

So the eigenvalues are $\lambda_1 = 1$ and $\lambda_2 = 0.8$. For $\lambda_1$,

\begin{align*}
0.85x + 0.15y &= x \\
0.95x + 0.05y &= y
\end{align*}

These equations imply that $(0.5, 0.5)'$ is a corresponding eigenvector. For $\lambda_1$,

\begin{align*}
0.85x + 0.15y &= 0.8x \\
0.95x + 0.05y &= 0.8y
\end{align*}

These equations imply that $(0.25, 0.75)'$ is the corresponding eigenvector.  Double-checking, we see that $(0.25, 0.75)'$ is the stationary distribution:

$$
\begin{bmatrix} 0.25 \\ 0.75 \end{bmatrix}\begin{bmatrix} 0.85 & 0.15 \\ 0.05 & 0.95 \end{bmatrix} = \begin{bmatrix} 0.25 \\ 0.75 \end{bmatrix}
$$

The unconditional mean of labor endowment is $(0.25)(0.7) + (0.75)(1.1) = 1$.

```{r 3b}
Q <- matrix(c(0.85, 0.05, 0.15, 0.95), nrow = 2)

# Numerical estimation
x <- c(0.5, 0.5)
for (i in 1:100) x <- x %*% Q
x
```

\pagebreak

(c) Take the following as parameters: $\beta = 0.95, \gamma = 3, r = 0.03, w = 1.1$. Discretize the asset choice into a finite grid. Note that this directly restricts the assets to lie in a compact set. Numerically solve for the value functions by iterating on the Bellman equation.

```{r 3c, echo = FALSE, eval = FALSE}
# parameters
beta <- .95
gamma <- 3
r <- 0.03
w <- 1.1
l_low <- 0.7
l_high <- 1.1

# assets grid
min_a <- 0.01
max_a <- 5
inc_a <- min_a
a <- seq(min_a, max_a, by = inc_a)
n <- length(a)

# initialize value grid 
v_low <- rep(0, times=n)
v_high <- rep(0, times=n)

# initialize decision rule grid
decision_low <- rep(0, times=n)
decision_high <- rep(0, times=n)

# create consumption matrices
# (columns are different values of a; rows are values of a')
ones <- rep(1, times=n)
c_matrix_low <- w * l_low + (1 + r)*ones %*% t(a)  - a %*% t(ones)
c_matrix_high <- w * l_high + (1 + r)*ones %*% t(a)  - a %*% t(ones)

# create flow utility matrices and set negative consumption levels to low value
utility_matrix_low <- (c_matrix_low^(1-gamma))/(1-gamma)
utility_matrix_low[c_matrix_low < 0|!is.finite(utility_matrix_low)] <- -1000

utility_matrix_high <- (c_matrix_high^(1-gamma))/(1-gamma)
utility_matrix_high[c_matrix_high < 0|!is.finite(utility_matrix_high)] <- -1000

# solve bellman equations:
test <- 10
iter <- 1
max_iter <- 10000

while (test != 0 & iter < max_iter) {
  # create value matrix for different a and a' values with transition probabilities 
  # to next period shock
  value_matrix_low <- utility_matrix_low + beta * Q[1, 1] * v_low %*% t(ones) + 
    beta * Q[1, 2] * v_high %*% t(ones)
  value_matrix_high <- utility_matrix_high + beta * Q[2, 1] * v_low %*% t(ones) + 
    beta * Q[2, 2] * v_high %*% t(ones)
  
  # find max utility
  tv_low <- apply(value_matrix_low, 2, max)
  tv_high <- apply(value_matrix_high, 2, max)
  
  # find column with max utility
  tdecision_low <- apply(value_matrix_low, 2, which.max)
  tdecision_high <- apply(value_matrix_high, 2, which.max)

  # update test variable so loop ends if no changes from previous iteration
  test <- max(tdecision_low - decision_low) + max(tdecision_high - decision_high)
  
  # update value grid
  v_low <- tv_low
  v_high <- tv_high

  # update decision rule
  decision_low <- tdecision_low
  decision_high <- tdecision_high
  
  iter <- iter + 1
}

# save policy function based on decision rule
policy_function_low <- tibble(a = a, a_prime = min_a + decision_low*inc_a) %>%
  mutate(c = w * l_low + (1 + r)*a  - a_prime)

policy_function_high <- tibble(a = a, a_prime = min_a + decision_high*inc_a) %>%
  mutate(c = w * l_high + (1 + r)*a  - a_prime)
```

i. Plot the value functions for the high and low values of the labor shocks. Do they appear to have the properties that we have assumed?

Yes, the value function appears continuous, increasing, and differentiable.  In addition, the value function for the high labor shocks is larger than the value function for low labor shocks.  The difference decreases as the level of current assets increases.

```{r 3ci, echo = FALSE, eval = FALSE}
plot(1, type = "n", xlim = c(min_a, max_a), ylim = c(-11, -6), 
     xlab = "Current Assets (a)", ylab = "Utility", main = "Value Function")

lines(x = a, y =v_high, col = "red")
lines(x = a, y =v_low, col = "blue")

text(x = 1, y= -7, "High Value of Labor Shock", col = "red")
text(x = 4, y= -8, "Low Value of Labor Shock", col = "blue")
```

\pagebreak

ii. Plot the asset holdings decision rules $a'(a, l_i)$ for the two labor shocks. Is there a value $\bar{a}$ such that $a'(a, l_2) < a$ for all $a \ge \bar{a}$? What do the features of the decision rules suggest about the long-run behavior of assets $a_t$?

As the chart shows at $\bar{a} \approx 2.2575$, $a'(a, l_2) < a$ for all $a \ge \bar{a}$.

The decision rules suggest that when a high labor shock is realized, agents save more than their current asset holdings and when a low labor shock is realized, agents draw down their current asset holdings.  In the long run, the transition matrix implies where along the range from 0 to $\bar{a}$, agents converge to.

```{r 3cii, echo = FALSE, eval = FALSE}

plot(1, type = "n", xlim = c(min_a, max_a), ylim = c(min_a, max_a), 
     xlab = "Current Assets (a)", ylab = "Future Assets (a')")

lines(x = policy_function_high$a, y =policy_function_high$a_prime, col = "red")
lines(x = policy_function_low$a, y =policy_function_low$a_prime, col = "blue")

text(x = 1, y= 2.5, "High Value of Labor Shock", col = "red")
text(x = 4, y= 2, "Low Value of Labor Shock", col = "blue")
text(x = 3.5, y= 4, "a' = a")

intersection_low <- policy_function_low %>% 
  filter(a == a_prime) %>%
  summarise_all(mean)

intersection_high <- policy_function_high %>% 
  filter(a == a_prime) %>%
  summarise_all(mean)

points(x = intersection_low$a, y = intersection_low$a_prime, pch=16, col = "blue")
points(x = intersection_high$a, y = intersection_high$a_prime, pch=16, col = "red")

abline(a = 0, b = 1, lty = 2)
```

\pagebreak

(d) To find the stationary distribution over assets and labor, first construct a (large) transition matrix over $(a, l)$. Then starting from an arbitrary initial distribution, iterate until convergence on the mapping from the current distribution into the distribution in the next period. What is the mean of the stationary asset distribution? Plot the (marginal) asset distribution.

```{r 3d, echo = FALSE, eval = FALSE}
tol <- .05
diff <- 1000

p_low0 <- matrix(rep(1/n, times = 2*n), ncol = 2)
p_high0 <- matrix(rep(1/n, times = 2*n), ncol = 2)

p_low <- p_low0
p_high <- p_high0

iter <- 1

while (diff > tol & iter < max_iter ) {

  p_low <- 0 * p_low0
  p_high <- 0 * p_high0
  
  for (i in 1:n) {
    if (p_high0[i] > .Machine$double.xmin) {
      p_high[decision_high[i], 1] <- p_high[decision_high[i]] + p_high0[i, 1] * Q[2, 1]
      p_high[decision_high[i], 2] <- p_high[decision_high[i]] + p_high0[i, 2] * Q[2, 2]
    }
    
    if (p_low0[i] > .Machine$double.xmin) {
      p_low[decision_low[i], 1] <- p_low[decision_low[i]] + p_low0[i, 1] * Q[1, 1]
      p_low[decision_low[i], 2] <- p_low[decision_low[i]] + p_low0[i, 2] * Q[1, 2]
    }
    
  }
  
  diff <- sum(abs(p_low - p_low0) + abs(p_high - p_high0))
  
  p_low0 <- p_low
  p_high0 <- p_high
  
  iter <- iter + 1
}

p_low_marginal <- apply(p_low, 1, sum) / sum(p_low)
p_high_marginal <- apply(p_high, 1, sum) / sum(p_high)

mean_assets_low <- sum(a * p_low_marginal)
mean_assets_high <- sum(a * p_high_marginal)
mean_assets <- (mean_assets_low + mean_assets_high)/2
```

```{r 3d_plot, echo = FALSE, eval = FALSE}
plot(1, type = "n", xlim = c(min_a, max_a), ylim = c(0, 1), main = "Stationary Asset CDF", xlab = "Current Assets (a)", ylab = "CDF")

lines(x = a, y = cumsum(p_low_marginal), col = "blue")
lines(x = a, y = cumsum(p_high_marginal), col = "red")

abline(v=mean_assets, lty=2)
abline(v=mean_assets_low, col = "blue", lty=2)
abline(v=mean_assets_high, col = "red", lty=2)

```

\pagebreak

## Appendix

```{r appendix, ref.label='3c', eval = FALSE}
```

```{r appendix, ref.label='3d', eval = FALSE}
```