---
title: "ECON 703 - PS 4"
author: "Alex von Hafften^[I worked on this problem set with a study group of Michael Nattinger, Andrew Smith, and Ryan Mather. I also discussed problems with Emily Case, Sarah Bass, and Danny Edgel.]"
date: "9/9/2020"
output: pdf_document
header-includes:
- \newcommand{\N}{\mathbb{N}}
- \newcommand{\Z}{\mathbb{Z}}
- \newcommand{\R}{\mathbb{R}}
- \newcommand{\Q}{\mathbb{Q}}
- \usepackage{bm}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

(1) Let $X, Y$ be two vector spaces such that dim $X = n$, dim $Y = m$.  Construct a basis of $L(X, Y)$.

Let $B_X = \{u_1, ..., u_n\}$ be a basis for $X$. Consider $x \in X$ such that $x=c_1u_1+...+c_nu_n$ for $c_1, ..., c_n\in \R$. Let $B_Y = \{v_1, ..., v_m\}$ be a basis for $Y$, so for any $T \in L(X, Y)$, $T(u_i)=b_{i1}v_1+...+b_{im}v_m$ for $i \in \{1, ..., n\}, b_{i1}, ..., b_{im}\in \R$.  Thus,

\begin{align*}
T(x) &= T(c_1u_1+...+c_nu_n) \\
&= c_1T(u_1)+...+c_nT(u_n) \\
&= c_1(b_{11}v_1+...+b_{1m}v_m)+...+c_n(b_{n1}v_1+...+b_{nm}v_m) \\
&= c_1b_{11}v_1+...+c_1b_{1m}v_m+...+c_nb_{n1}v_1+...+c_nb_{nm}v_m \\
&= \begin{pmatrix} v_1 & ... & v_m \end{pmatrix} \begin{pmatrix} b_{11} & ... & b_{n1} \\ \vdots &  & \vdots \\ b_{1m} & ... & b_{nm} \end{pmatrix} \begin{pmatrix} c_1 \\ \vdots \\ c_n \end{pmatrix}  \\
\end{align*}

Therefore, $B_{L(X, Y)}$ forms a basis for $L(X, Y)$:

\begin{align*}
B_{L(X, Y)}
&= \Bigg\{ \begin{pmatrix} 1 & ... & 0 \\ \vdots &  & \vdots \\ 0 & ... & 0 \end{pmatrix}_{n\times m}, ..., \begin{pmatrix} 0 & ... & 0 \\ \vdots &  & \vdots \\ 0 & ... & 1 \end{pmatrix}_{n\times m}\Bigg\} \\
&=\{ A_{n\times_m} | a_{ij} = 1 \text{ and }a_{lk}=0 \forall i \in \{1,..., n\},j \in\{1, ..., m\}, l\neq i, k \neq j \} 
\end{align*}

$B_{L(X, Y)}$ spans $L(X, Y)$ where $b_{ij}$ is the coefficient on the element of $B_{L(X, Y)}$ with a one in the $i$th column and the $j$th row.  $B_{L(X, Y)}$ is also linear independent because each element has a single nonzero cell whose location is to unique to the nonzero cells in other elements. Since no element of $B_{L(X, Y)}$ has the same nonzero cell as another element, any linear combination of the elements that equals an $n \times m$ matrix of zeros implies that all coefficients on elements of $B_{L(X, Y)}$ are zero.

\pagebreak

(2) Suppose that $T \in L(X, X)$ and $\lambda$ is $T$'s eigenvalue.

(a) Prove that $\lambda^k$ is an eigenvalue of $T^k$, $k\in \N$.

Proof: If $\lambda$ is $T$'s eigenvalue $\implies T(v)=\lambda v$ for eigenvector $v \neq \bar{0}$.  Applying $T$ again, we get $T(T(v)) = \lambda(\lambda v) \implies T^2(v)=\lambda^2v$.  Similarly, applying $T$ $k$ times to $v$, we get $T^k(v) = \lambda^k v$.  Thus, $\lambda^k$ is an eigenvalue for $T^k$ where $k \in \N$.

(b) Prove that if $T$ is invertible, then $\lambda^{-1}$ is an eigenvalue of $T^{-1}$.

Proof: If $T$ is invertible, $T^{-1} \in L(X, X)$. Thus, for eigvenvector $v$ of $T$,

\begin{align*}
T(v) = \lambda v 
&\implies T^{-1}(T(v))=T^{-1}(\lambda v) \\ 
&\implies v=\lambda T^{-1}(v) \\
&\implies T^{-1}(v) = \lambda^{-1}v.
\end{align*}

Therefore, $\lambda^{-1}$ is an eigenvalue of $T^{-1}$.

(c) Define an operator $S:X \to X$, such that $S(x) = T(x) - \lambda x$ for all $x \in X$. Is $S$ linear? Prove that ker $S :=\{x \in X | S(x) = \bar{0}\}$ is a vector space.

For $x_1, x_2 \in X$ and $\alpha_1, \alpha_2 \in \R$,

\begin{align*}
S(\alpha_1 x_1 + \alpha_2 x_2) &= T(\alpha_1 x_1 + \alpha_2 x_2) - \lambda(\alpha_1 x_1 + \alpha_2 x_2) \\
&= \alpha_1 T(x_1) + \alpha_2 T(x_2) - \lambda\alpha_1 x_1 - \lambda\alpha_2 x_2\\
&= \alpha_1 (T(x_1) - \lambda x_1)  + \alpha_2 (T(x_2) - \lambda x_2)\\
&= \alpha_1 S(x_1) + \alpha_2 S(x_2)
\end{align*}

Yes, $S$ is linear.

Proof: For a fixed $\lambda$, let $x, y \in$ ker $S$ and $\alpha, \beta \in \R$. We know that $S(x) = S(y) = \bar{0}$. As a linear transformation, $S(\alpha x + \beta y) = \alpha S(x) + \beta S(y) =\alpha \bar{0} + \beta \bar{0} = \bar{0}$, so properties 1, 2, 5, 6, 7 of the definition of a vector space are satisfied.

For property 3, note that $S(\bar{0}) = T(\bar{0})-\lambda \bar{0}=T(\bar{0})=\bar{0}$, so $\bar{0} \in$ ker$S$.\footnote{For any linear transformation $T:X\to Y$, $T(\bar{0}_X) = T(\bar{0}_X+\bar{0}_X) = T(\bar{0}_X) + T(\bar{0}_X) \implies T(\bar{0}_X)=T(\bar{0}_X)-T(\bar{0}_X)=\bar{0}_Y$.} Furthermore, $x + \bar{0} = \bar{0} + x = x$ for $x \in$ ker$S$.

For property 4, for $x \in$ ker$S$, $S(-x)=(-1)S(x) =(-1)\bar{0}=\bar{0}$ where $x+(-x)=\bar{0}$.

For property 8, for $x \in$ ker$S$, $S(1 \cdot x) = 1\cdot S(x) = 1\cdot \bar{0}=\bar{0}$.

Thus ker$S$ is a vector space. $\square$

\pagebreak

(3) Let $T : \R^2 \to \R^2$ be given by $T(x, y) = (x-y, 2x+3y)$.  Let $W$ be the standard basis of $\R^2$ and let $V$ be another basis of $\R^2, V = \{ (1, -4), (-2, 7)\}$ in the coordinates of $W$.

(a) Find mtx$_W(T)$.

\begin{align*}
T(x, y) &= (x-y)w_1+(2x+3y)w_2 \\
&= (w_1+2w_2)x +(-w_1+3w_2)y
\end{align*}

mtx$_W(T) = \begin{pmatrix} 1 & -1 \\ 2 & 3 \end{pmatrix}$.

(b) Find mtx$_V(T)$.

$P =$ mtx$_{W,V}(id)=\begin{pmatrix} 1 & -2 \\ -4 & 7 \end{pmatrix}$

$P^{-1} = \begin{pmatrix} -7 & -2 \\ -4 & -1 \end{pmatrix}$

mtx$_V(T) = P^{-1}$mtx$_W(T)P = \begin{pmatrix} -7 & -2 \\ -4 & -1 \end{pmatrix}\begin{pmatrix} 1 & -1 \\ 2 & 3 \end{pmatrix}\begin{pmatrix} 1 & -2 \\ -4 & 7 \end{pmatrix} = \begin{pmatrix} -15 & 29 \\ -10 & 19 \end{pmatrix}$

(c) Find $T(1, -2)$ in the basis $V$.

mtx$_V(T)$mtx$_{V,W}(id) \begin{pmatrix} 1 \\ -2 \end{pmatrix} = \begin{pmatrix} -15 & -10 \\ 29 & 19 \end{pmatrix}\begin{pmatrix} -7 & -2 \\ -4 & -1 \end{pmatrix}\begin{pmatrix} 1 \\ -2 \end{pmatrix} = \begin{pmatrix} -15 & -10 \\ 29 & 19 \end{pmatrix}\begin{pmatrix} -3 \\ -2 \end{pmatrix}= \begin{pmatrix} -13 \\ -8 \end{pmatrix}$

Check: $T(1, -2) = (1-(-2), 2(1)+3(-2)) = (3, -4)$ and $(-13)(1, -4)+(-8)(-2, 7)=(-13, 52)+(16, -56)=(3, -4)$.

\pagebreak

(4) In this exercise you will learn to solve first order linear difference equations in $n$ variables.  We want to find an $n$-dimensional process $\{\mathbf{x}_1, \mathbf{x}_2, ...\}$ such that each $\mathbf{x}_i$ is an $n$-dimensional vector and 

\begin{align}
\mathbf{x}_t = A\mathbf{x}_{t-1}, t=1, 2,...,
\end{align}

|         where $A \in M_{n \times n}$ and $\mathbf{x}_0 \in \R^n$ are given. Then

$$
\mathbf{x}_1 = A \mathbf{x}_0, \mathbf{x}_2 = A \mathbf{x}_1 = A(A \mathbf{x}_0)= A^2 \mathbf{x}_0, \mathbf{x}_t = A^t \mathbf{x}_0 \forall t \in \N,
$$

|         where $A^t = A \cdot A \cdot ... \cdot A$ ($t$ times). Thus, we need to calculate $A^t$.  
|         To do this, we diagonalize $A$, $A = PDP^{-1}$, where $D$ is diagonal, $D = diag\{\lambda_1, ..., \lambda_n\}$. 
|         Hence we can rewrite

$$
A^t = PDP^{-1}PDP^{-1}...PDP^{-1}=PD^tP^{-1}=Pdiag\{\lambda^t_1, ..., \lambda^t_n\}P^{-1},
$$

|         which is now easy to compute. Thus, what you is

|         Step 1: Calculate $A$'s eigenvalues $\lambda_1,...,\lambda_n$ and eigenvectors $\mathbf{v}_1, ..., \mathbf{v}_n$.  
|         Remember that we need to independent eigenvectors (this holds if all eigenvalues are distinct).

|         Step 2: Set $D =diag\{\lambda_1, ..., \lambda_n\}$ and $P = \{\mathbf{v}_1, ..., \mathbf{v}_n\}$ (eigenvectors are columns of $P$).

|         Step 3: Calculate $P^{-1}$ and $Pdiag\{\lambda_1^t, ..., \lambda_n^t\}P^{-1}$.

|         Step 4: Plug $A^t$ from Step 3 to get $\mathbf{x}_t = A^t\mathbf{x}_0$.

|         Implement the above approach to solve for $\mathbf{x}_t \in \R^2$:

$$
\mathbf{x}_t = \begin{pmatrix} 1 & 4 \\ 2 & -1 \end{pmatrix}\mathbf{x}_{t-1}, \mathbf{x}_0 = \begin{pmatrix} 1 \\ 1 \end{pmatrix}
$$

|         Simplify your answer as much as possible.

The characteristic polynomial is $(1-\lambda)(-1-\lambda)-4(2)=\lambda^2-9=(\lambda-3)(\lambda+3)$, so the eigenvalues are $\lambda_1 = 3$ and $\lambda_2 = -3$. The eigenvectors are thus solutions to:

\begin{align*}
\begin{pmatrix} -2 & 4 \\ 2 & -4 \end{pmatrix}\mathbf{v}_1 &= \bar{0} \\
\begin{pmatrix} 4 & 4 \\ 2 & 2 \end{pmatrix}\mathbf{v}_2 &= \bar{0}
\end{align*}

Thus, the eigenvectors are $\Bigg\{ \begin{pmatrix} 2 \\ 1 \end{pmatrix}, \begin{pmatrix} -1 \\ 1\end{pmatrix}\Bigg\}$. Thus,

\begin{align*}
D &= \begin{pmatrix} 3 & 0 \\ 0 & -3 \end{pmatrix} \\
P &= \begin{pmatrix} 2 & -1 \\ 1 & 1 \end{pmatrix} \\
P^{-1} &= \begin{pmatrix} 1/3 & 1/3 \\ -1/3 & 2/3 \end{pmatrix}
\end{align*}

Thus,

\begin{align*}
\mathbf{x}_t &= PD^tP^{-1}\mathbf{x}_0 \\
&= \begin{pmatrix} 2 & -1 \\ 1 & 1 \end{pmatrix} \begin{pmatrix} 3^t & 0 \\ 0 & (-3)^t \end{pmatrix} 
\begin{pmatrix} 1/3 & 1/3 \\ -1/3 & 2/3 \end{pmatrix} 
\begin{pmatrix} 1 \\ 1 \end{pmatrix} \\
&= \begin{pmatrix} 2 & -1 \\ 1 & 1 \end{pmatrix} 
\begin{pmatrix} 3^t & 0 \\ 0 & (-3)^t \end{pmatrix} 
\begin{pmatrix} 2/3 \\ 1/3 \end{pmatrix} \\
&= \begin{pmatrix} 2 & -1 \\ 1 & 1 \end{pmatrix} 
\begin{pmatrix} 2*3^{t-1} \\ -(-3)^{t-1} \end{pmatrix} \\
&= \begin{pmatrix} 4*3^{t-1} + (-3)^{t-1} \\ 2*3^{t-1} - (-3)^{t-1} \end{pmatrix}
\end{align*}

Below is R code verifying the answer:

```{r}
library(matlib)

A <- matrix(c(1, 2, 4, -1), ncol=2)
print(A)
ev <- eigen(A)
p <- t(t(ev$vectors))
d <- diag(ev$values)
print(p)
print(d)
print(p %*% d %*% inv(p))

# for first ten x_t
x_0 <- c(1, 1)

for (t in 1:10) {
  print(paste("t =", t))
  print(p %*% d^t %*% inv(p) %*% x_0)
}
```

\pagebreak

(5) In this exercise you will learn to to solve $n$th order linear difference equations in one variable.  We want to find a sequence of real numbers $\{z_t\}_{t=1}^\infty$, which satisfies

\begin{align}
z_t = a_1 z_{t-1} + a_2 z_{t-2} + ... + a_n z_{t-n},
\end{align}

|         where $a_1, ..., a_n \in \R$ and $z_0, z_{-1}, ..., z_{-n+1} \in \R$ are given.

(a) Define $\mathbf{x}_t := (z_t,z_{t-1},...,z_{t-n+1})'$ and rewrite Eq. (2) in the form of Eq. (1). What is $A$?

$$A_{n \times n} = \begin{pmatrix} a_1 & a_2 & ... & a_{n-1} & a_n \\ 1 & 0 & ... & 0 & 0 \\ 0 & 1 & ... & 0 & 0 \\ \vdots &  \vdots & & \vdots & \vdots \\ 0 & 0 & ... & 1 & 0 \end{pmatrix}$$

\begin{align*}
A\mathbf{x}_{t-1}
&= \begin{pmatrix} a_1 & a_2 & a_3 & ... & a_{n-2} & a_{n-1} & a_n \\ 1 & 0 & 0 & ... & 0 & 0 & 0 \\ 0 & 1 & 0 & ... & 0 & 0 & 0 \\ \vdots & \vdots & \vdots & & \vdots & \vdots & \vdots \\ 0 & 0 & 0 & ... & 1 & 0 & 0 \\ 0 & 0 & 0 & ... & 0 & 1 & 0 \end{pmatrix} \begin{pmatrix} z_{t-1} \\ z_{t-2} \\ z_{t-3} \\ \vdots \\ z_{t-n+1} \\ z_{t-n} \end{pmatrix} \\
&= \begin{pmatrix} a_1 z_{t-1} + a_2 z_{t-2} + ... + a_{n-1} z_{t-n+1} + a_n z_{t-n} \\ z_{t-1} \\ z_{t-2} \\ \vdots \\ z_{t-n+1} \end{pmatrix}\\
&= \begin{pmatrix} z_t \\ z_{t-1} \\ z_{t-2} \\ \vdots \\ z_{t-n+1} \end{pmatrix}\\
&= \mathbf{x}_{t}
\end{align*}

\pagebreak

(b) Notice that if you find the function form of $z_t = f(t)$, then you do not need to find a similar form for $z_{t-1},...,z_{t-n+1}$ (you use the same function $f(\cdot)$ and evaluate it at a different time).  Thus, you actually do not need to calculate $Pdiag\{\lambda_1^t, ..., \lambda_n^t\}P^{-1}\mathbf{x}_0$. You only need the first coordinate of that $n$-dimensional vector.  The first coordinate takes the form

\begin{align}
\mathbf{x}_{t1} \equiv z_t = c_1\lambda_1^t + c_2\lambda_2^t + ... + c_n \lambda_n^t,
\end{align}

|         where coefficient $c_1, ..., c_n$ depend on $P$ and $\mathbf{x}_0$.  Given Eq. (3) which holds for any $t$ and initial values 
|         $z_0, ..., z_{-n+1}$, which equations must $c_1, ..., c_n$ solve?

\begin{align*}
\mathbf{x}_{0} 
= \begin{pmatrix} z_0 \\ z_{-1} \\ \vdots \\ z_{-n+1} \end{pmatrix}
&= \begin{pmatrix} c_1\lambda_1^0 + c_2\lambda_2^0 + ... + c_n\lambda_n^0 \\ c_1\lambda_1^{-1} + c_2\lambda_2^{-1} + ... + c_n\lambda_n^{-1} \\ \vdots \\ c_1\lambda_1^{-n+1} + c_2\lambda_2^{-n+1} + ... + c_n\lambda_n^{-n+1} \end{pmatrix} \\ 
\implies 
\begin{pmatrix} z_0 \\ z_{-1} \\ \vdots \\ z_{-n+1} \end{pmatrix}
&= \begin{pmatrix} 1 & 1 & ... & 1 \\ \lambda_1^{-1} & \lambda_2^{-1} & ... & \lambda_n^{-1} \\ \vdots & & & \vdots \\ \lambda_1^{-n+1} & \lambda_2^{-n+1} & ... & \lambda_n^{-n+1} \end{pmatrix} \begin{pmatrix} c_1 \\ c_2 \\ \vdots \\ c_{n} \end{pmatrix} \\
\end{align*}

(c) Suppose that $n=3$, $a_1 = 2$, $a_2 = 1$, $a_3 = -2$, and $z_0 = 2$, $z_{-1}=2$, $z_{-2}=1$.  Find the expression for $z_t$ as a function of $t$.

$$
A = \begin{pmatrix} 2 & 1 & -2 \\ 1 & 0 & 0 \\ 0 & 1 & 0\end{pmatrix}
$$

\begin{align*}
\det(A - \lambda I) &= 0 \\
\implies 
\det \begin{pmatrix} 2-\lambda & 1 & -2 \\ 1 & -\lambda & 0 \\ 0 & 1 & -\lambda \end{pmatrix} &= 0 \\
\implies 
(2-\lambda)(-\lambda)(-\lambda)+(-2)(1)(1)-(1)(1)(-\lambda)&=0 \\
\implies 
-\lambda^3 + 2 \lambda^2 + \lambda - 2 &=0 \\
\end{align*}

Roots at $\lambda_1 = -1, \lambda_2 = 1, \lambda_3 = 2$. Thus,

\begin{align*}
\begin{pmatrix} 
2 \\ 2 \\ 1
\end{pmatrix} &= 
\begin{pmatrix} 
1 & 1 & 1 \\ 
(-1)^{-1} & (1)^{-1} & (2)^{-1} \\ 
(-1)^{-2} & (1)^{-2} & (2)^{-2}
\end{pmatrix}
\begin{pmatrix} 
c_1 \\ c_2 \\ c_3
\end{pmatrix} \\ 
\begin{pmatrix} 
2 \\ 2 \\ 1
\end{pmatrix} &= 
\begin{pmatrix} 
c_1 + c_2 +c_3 \\ -c_1 +c_2 + (1/2)c_3 \\ c_1 + c_2 + (1/4)c_3
\end{pmatrix}
\end{align*}

Thus, $c_1 = -1/3, c_2 = 1, c_3 = 4/3 \implies z_t = (-1)^t(-1/3)+1+2^t(4/3)$.
