---
title: "ECON 711B - Voting Game"
author: "Alex von Hafften"
date: "11/5/2020"
output: pdf_document
header-includes:
- \newcommand{\N}{\mathbb{N}}
- \newcommand{\Z}{\mathbb{Z}}
- \newcommand{\R}{\mathbb{R}}
- \newcommand{\Q}{\mathbb{Q}}
- \newcommand{\Lf}{\mathcal{L}}
- \usepackage{bm}
- \usepackage{bbm}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Ideas for setup:

- Continuum of players.

- Discrete choices: Vote and not vote. Democratic vote for democrats; republicans vote for republicans.

Stylized facts:

- Costly voting.

- Signal from voting increases with income.

Result:

- Voting rates increase with income.


## Higher income and voting

- Two types of players: R and D.

- If more Rs vote than Ds, Rs win and Rs get payoff $x$.  If more Ds vote than Rs, Ds win and Ds get payoff $x \in \R_+$.

- There is a continuum of each type of players.  For each type, player $w \in [0,1]$ earns $w$ per unit of time.

- It takes $c \in (0,1)$ units of time to vote.  So the cost of voting is $cw$.

- The benefit of voting is a signal of your in-group-ness.  The strength of the signal is $f(w)$. Assume that $f$ is strictly increasing ($f' > 0$) and strictly convex ($f'' > 0$).

- If no one is voting, everyone has an incentive to vote because their party would win.

- The payoff for a D is

+------+---------+-----------+
|$u_D=$| R wins  | D wins    |
+------+---------+-----------+
|$v=1$ |$f(w)-cw$|$f(w)-cw+x$|
+------+---------+-----------+
|$v=0$ | $0$     |$x$        |
+------+---------+-----------+

- The payoff for a R is

+------+---------+-----------+
|$u_R=$| D wins  | R wins    |
+------+---------+-----------+
|$v=1$ |$f(w)-cw$|$f(w)-cw+x$|
+------+---------+-----------+
|$v=0$ | $0$     |$x$        |
+------+---------+-----------+

- Since the unit mass of an individual is zero, they cannot sway the result of the election.

- Voting is a dominant strategy if $cw < f(w)$.

- Not voting is a dominant strategy if $cw > f(w)$

```{r, echo = FALSE}

c <- 0.4

x <- seq(0, 1, .01)

v_p <- x^2-c*x

nv_p <- rep(0, times = 101)

plot(1, type = "n", xlim=c(0,1), ylim = c(min(v_p), max(v_p)), xlab = "w", ylab = "Payoff")

lines(x=x, y=v_p, col = "blue")
text("u(v=1)", x = 0.8, y=0.4, pos = 2, col ="blue")

lines(x=x, y=nv_p, col = "red")
text("u(v=0)", x = 1, y=0.1, pos = 2, col="red")

abline(v= c)
text("f(w) = cw", x = c, y=0.4, pos = 2)
```

- Result is lower income agents don't vote and higher income agents vote despite the higher time-cost trade-off.

