---
title: "ECON 711 - PS 2"
author: "Alex von Hafften^[I worked on this problem set with a study group of Michael Nattinger, Andrew Smith, Tyler Welch, and Ryan Mather. I also discussed problems with Emily Case, Sarah Bass, and Danny Edgel.]"
date: "9/21/2020"
output: pdf_document
header-includes:
- \newcommand{\N}{\mathbb{N}}
- \newcommand{\Z}{\mathbb{Z}}
- \newcommand{\R}{\mathbb{R}}
- \newcommand{\Q}{\mathbb{Q}}
- \newcommand{\bd}{\text{ bd}}
- \usepackage{bm}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Question 1. Convex production sets, concave production functions, convex costs

Consider a production function $f : \R^m_+ \to \R_+$ for a single-output firm.

(a) Prove that if the production set $Y = \{(q, -z) : f(z) \ge q\} \subset \R^{m+1}$ is convex, the production function $f$ is concave.

Proof: Choose $(q, -z), (q', -z') \in Y$ such that $f(z)=q$ and $f(z')=q'$. The convexity of $Y$ implies that $t(q, -z) + (1-t) (q', -z') \in Y$ for $t \in (0, 1)$. Thus, $f(tz + (1-t)z') \ge tq+(1-t)q'$ by the definition of $Y$. Our choice of $(q, -z), (q', -z') \implies f(tz + (1-t)z') \ge tf(z)+(1-t)f(z')$.  Therefore, $f$ is concave. $\square$

(b) Prove that if $f$ concave, the cost function

$$
c(q, w) = \min w \cdot z \text{ subject to } f(z) \ge q
$$

is convex in $q$.

Proof:  Fixing $w \in \R^k_+$, choose $q, q' \in \R$.  Define $z \in Z^*(q, w)$, $z' \in Z^*(q', w)$, and $\tilde{z} \in Z^*(tq+(1-t)q', w)$ for $t \in (0, 1)$.  By the concavity of $f$,

\begin{align*}
\tilde{z} &\le tz+(1-t)z' \\
\implies w\tilde{z} &\le w(tz+(1-t)z') \\
\implies w\tilde{z} &\le twz+(1-t)wz') \\
\implies c(f(\tilde{z}), w) &\le tc(f(z), w)+(1-t)c(f(z'), w) \\
\implies c(tq+(1-t)q', w) &\le tc(q, w)+(1-t)c(q', w)
\end{align*}

Therefore, $c$ is convex. $\square$

\pagebreak

# Question 2. Solving for the profit function given technology...

Let $k=2$, and let the production set be

$$
Y = \{(y_1, y_2) : y_1 \le 0 \text{ and } y_2 \le B(-y_1)^{\frac{2}{3}}\}
$$

where $B > 0$ is a known constant.  Assume both prices are strictly positive.

(a) Draw $Y$, or describe it clearly.

```{r, echo=FALSE}
plot(1, type="n", ylim=c(-10, 10), xlim=c(-10, 10), xaxs="i", yaxs="i", xlab="y_1", ylab="y_2", axes = FALSE)

B <- 2

y_1 <- ((-10*100):0)/100
y_2 <- 2*(-y_1)^(2/3)

polygon(x = c(y_1, 0, -10, -10), y=c(y_2, -10, -10, 10), col="red", border = "red")

lines(x=y_1, y=y_2)

abline(h=0, v=0)

abline(v=-1, h=B, lty=2)
text("y_2 = B", x=5, y=3)
text("y_1 = -1", x=-2.5, y=7)

box()

```

(b) Solve the firm's profit maximization problem to find $\pi(p)$ and $Y^*(p)$.

The firm's profit is

$$
\pi(p) = \max_{y_1, y_2 \in Y}\{p_1y_1+p_2y_2\}
$$

Define $z = -y_1$.  Notice that the firm will produce $y_2 = Bz^{2/3}$ because it is the maximum output given $z$ units of input.  Thus, we can rewrite the firm's profit function as 

$$
\pi(p) = \max_{z}\{p_1(-z)+p_2Bz^{2/3}\} = \max_{z}\{p_2Bz^{2/3}-p_1z\}
$$

Setting the first order condition of the profit function to zero:

\begin{align*}
\frac{\partial\pi}{\partial z} &= p_2(2/3)Bz^{-1/3}-p_1 \\
z^* &= \Bigg(\frac{2p_2B}{3p_1}\Bigg)^3 \\
\end{align*}

Plugging $z^*$ into transformations for $y_1, y_2$:

\begin{align*}
y^*_1 &= -\Bigg(\frac{2p_2B}{3p_1}\Bigg)^3 \\
y^*_2 &= B\Bigg(\Bigg(\frac{2p_2B}{3p_1}\Bigg)^3\Bigg)^{2/3} \\
 &= B^3\Bigg(\frac{2p_2}{3p_1}\Bigg)^2 \\
\end{align*}

Notice that $Y^*(p)$ is single-valued:

\begin{align*}
Y^*(p) = \Bigg\{(y_1, y_2) : y_1 = -\Bigg(\frac{2p_2B}{3p_1}\Bigg)^3, y_2 = B^3\Bigg(\frac{2p_2}{3p_1}\Bigg)^2 \Bigg \} \implies 
y(p) = \Bigg(-\Bigg(\frac{2p_2B}{3p_1}\Bigg)^3, B^3\Bigg(\frac{2p_2}{3p_1}\Bigg)^2\Bigg)
\end{align*}

Thus, the profit function is

\begin{align*}
\pi(p) &= p_1\Bigg(-\Bigg(\frac{2p_2B}{3p_1}\Bigg)^3\Bigg) + p_2\Bigg( B^3\Bigg(\frac{2p_2}{3p_1}\Bigg)^2\Bigg)\\
&= \frac{B^3p_2^3}{p_1^2}\Bigg(\frac{2^2}{3^2}-\frac{2^3}{3^3}\Bigg) \\
&= \frac{4B^3p_2^3}{27p_1^2} \\
\end{align*}


(c) Since $Y^*(p)$ is single-valued, I'll refer to it below as $y(p)$.  Verify that $\pi(\cdot)$ is homogeneous of degree 1, and $y(\cdot)$ is homogeneous of degree 0.

For $\alpha \in \R$:

\begin{align*}
\pi(\alpha p) 
&= \frac{4B^3(\alpha p_2)^3}{27(\alpha p_1)^2} \\
&= \alpha \frac{4B^3(p_2)^3}{27(p_1)^2} \\
y(\alpha p) &= \Bigg(-\Bigg(\frac{2(\alpha p_2)B}{3(\alpha p_1)}\Bigg)^3, B^3\Bigg(\frac{2(\alpha p_2)}{3(\alpha p_1)}\Bigg)^2\Bigg) \\
&= \Bigg(-\Bigg(\frac{2p_2B}{3p_1}\Bigg)^3, B^3\Bigg(\frac{2p_2}{3p_1}\Bigg)^2\Bigg)
\end{align*}

\pagebreak

(d) Verify that $y_1(p) = \frac{\partial\pi}{\partial p_1}(p)$ and $y_2(p) = \frac{\partial\pi}{\partial p_2}(p)$.

\begin{align*}
\frac{\partial\pi}{\partial p_1}(p) 
&= \frac{4B^3p_2^3}{27p_1^3}(-2)\\
&= -\frac{8B^3p_2^3}{27p_1^3}\\
&= -\Bigg(\frac{2p_2B}{3p_1}\Bigg)^3\\
&= y_1(p) \\
\frac{\partial\pi}{\partial p_2}(p) &= \frac{4B^3p_2^2}{27p_1^2}(3)\\
&= \frac{4B^3p_2^2}{9p_1^2}\\
&= B^3\Bigg(\frac{2p_2}{3p_1}\Bigg)^2 \\
&= y_2(p)
\end{align*}

(e) Calculate $D_p y(p)$, and verify it is symmetric, positive semidefinite, and $[D_p y]p = 0$

\begin{align*}
D_p y(p) &= \begin{pmatrix} \frac{\partial y_1}{\partial p_1}(p) & \frac{\partial y_2}{\partial p_1}(p) \\ 
                            \frac{\partial y_1}{\partial p_2}(p) & \frac{\partial y_2}{\partial p_2}(p) \end{pmatrix} \\
         &= \begin{pmatrix} \frac{8 p_2^3 B^3}{9 p_1^4}  & \frac{-8 p_2^2 B^3}{9 p_1^3} \\ 
                            \frac{-8 p_2^2 B^3}{9 p_1^3} & \frac{8p_2B^3}{9p_1^2} \end{pmatrix} \\
\end{align*}

Since both off diagonal elements of $D_p y(p)$ equal $\frac{-8 p_2^2 B^3}{9 p_1^3}$, $D_p y(p)$ is symmetric.

$$B > 0, p_1 > 0, p_2 > 0 \implies \frac{8 p_2^3 B^3}{9 p_1^4} > 0$$

\begin{align*}
\det D_p y(p) &= \frac{8 p_2^3 B^3}{9 p_1^4}\frac{8p_2B^3}{9p_1^2} - \frac{-8 p_2^2 B^3}{9 p_1^3}\frac{-8 p_2^2 B^3}{9 p_1^3}\\
              &= \frac{64 p_2^4 B^6}{81 p_1^6} - \frac{64 p_2^4 B^6}{81 p_1^6}\\
              &= 0
\end{align*}

Therefore, $D_p y(p)$ is positive semidefinite.

\begin{align*}
[D_p y]p &=  \begin{pmatrix} \frac{8 p_2^3 B^3}{9 p_1^4}  & \frac{-8 p_2^2 B^3}{9 p_1^3} \\ 
                            \frac{-8 p_2^2 B^3}{9 p_1^3} & \frac{8p_2B^3}{9p_1^2} \end{pmatrix}
             \begin{pmatrix} p_1 \\ p_2 \end{pmatrix} =
             \begin{pmatrix} \frac{8 p_2^3 B^3}{9 p_1^3}  + \frac{-8 p_2^3 B^3}{9 p_1^3} \\ 
                            \frac{-8 p_2^2 B^3}{9 p_1^2}  + \frac{8p_2^2B^3}{9p_1^2} \end{pmatrix} =
             \begin{pmatrix} 0 \\ 0 \end{pmatrix} 
\end{align*}

\pagebreak

# Question 3 ...and recovering technology from the profit function

Finally, suppose we didn't know a firm's production set $Y$, but did know its profit function was $\pi(p) Ap_1^{-2}p_2^3$ for all $p_1, p_2 > 0$ and $A > 0$ a known constant.

(a) What conditions must hold for this profit function to be rationalizable?  (You don't need to check them.)

(b) Recall that the outer bound was defined as $Y^O = \{y : p \cdot y \le \pi(p) \text{ for all } p\in P\}$. In this case, this is $Y^O = \{(y_1, y_2) : p_1y_1+p_2y_2 \le Ap_1^{-2}p_2^3 \text{ for all } (p_1, p_2) \in \R^2_{++}\}$.  Show that any $y \in Y^O$ must have $y_1 \le 0$, i.e., that good 1 must be an input only.

(c) Dividing both sides by $p_2$ and moving $\frac{p_1}{p_2}y_1$ to the right-hand side, we can rewrite $Y^O$ as $Y^O=\{(y_1, y_2) : y_2 \le  Ap_1^{-2}p_2^3 - \frac{p_1}{p_2}y_1$ for all $(p_1, p_2) \in \R^2_{++}\}$. Since the expression on the right depends only on the price ratio $\frac{p_2}{p_1}$ rather than the two individual prices, we can let $r \equiv \frac{p_2}{p_1} > 0$ denote this ratio, and write $Y^O$ as $Y_O=\{(y_1, y_2) : y_2 \le  Ar^2 - \frac{y_1}{r}$ for all $r \in \R_{++}\}=\{(y_1, y_2) : y_2 \le  \min_{r>0}(Ar^2 - \frac{y_1}{r})\}$.  Solve this minimization problem, and describe the production set $Y^O$.

(d) Verify that a production set $Y$ equal to the set $Y^O$ you just calculated would generate the "data" $\pi(p)=Ap_1^{-2}p_2^3$ that we started with.  [Hint:$2^{-2/3}+2^{1/3}=\frac{27}{4}$].
