---
title: "ECON 711 - PS 2"
author: "Alex von Hafften^[I worked on this problem set with a study group of Michael Nattinger, Andrew Smith, Tyler Welch, and Ryan Mather. I also discussed problems with Emily Case, Sarah Bass, and Danny Edgel.]"
date: "9/21/2020"
output: pdf_document
header-includes:
- \newcommand{\N}{\mathbb{N}}
- \newcommand{\Z}{\mathbb{Z}}
- \newcommand{\R}{\mathbb{R}}
- \newcommand{\Q}{\mathbb{Q}}
- \newcommand{\bd}{\text{ bd}}
- \usepackage{bm}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Question 1. Convex production sets, concave production functions, convex costs

Consider a production function $f : \R^m_+ \to \R_+$ for a single-output firm.

(a) Prove that if the production set $Y = \{(q, -z) : f(z) \ge q\} \subset \R^{m+1}$ is convex, the production function $f$ is concave.

Proof: Choose $(q, -z), (q', -z') \in Y$ such that $f(z)=q$ and $f(z')=q'$. The convexity of $Y$ implies that $t(q, -z) + (1-t) (q', -z') \in Y$ for $t \in (0, 1)$. Thus, $f(tz + (1-t)z') \ge tq+(1-t)q'$ by the definition of $Y$. Our choice of $(q, -z), (q', -z') \implies f(tz + (1-t)z') \ge tf(z)+(1-t)f(z')$.  Therefore, $f$ is concave. $\square$

(b) Prove that if $f$ concave, the cost function

$$
c(q, w) = \min w \cdot z \text{ subject to } f(z) \ge q
$$

is convex in $q$.

Proof:  Fixing $w \in \R^k_+$, choose $q, q' \in \R$.  Define $z \in Z^*(q, w)$, $z' \in Z^*(q', w)$, and $\tilde{z} \in Z^*(tq+(1-t)q', w)$ for $t \in (0, 1)$.  By the concavity of $f$,

\begin{align*}
\tilde{z} &\le tz+(1-t)z' \\
\implies w\tilde{z} &\le w(tz+(1-t)z') \\
\implies w\tilde{z} &\le twz+(1-t)wz') \\
\implies c(f(\tilde{z}), w) &\le tc(f(z), w)+(1-t)c(f(z'), w) \\
\implies c(tq+(1-t)q', w) &\le tc(q, w)+(1-t)c(q', w)
\end{align*}

Therefore, $c$ is convex. $\square$

\pagebreak

# Question 2. Solving for the profit function given technology...

Let $k=2$, and let the production set be

$$
Y = \{(y_1, y_2) : y_1 \le 0 \text{ and } y_2 \le B(-y_1)^{\frac{2}{3}}\}
$$

where $B > 0$ is a known constant.  Assume both prices are strictly positive.

(a) Draw $Y$, or describe it clearly.

```{r, echo=FALSE}
plot(1, type="n", ylim=c(-10, 10), xlim=c(-10, 10), xaxs="i", yaxs="i", xlab="y_1", ylab="y_2", axes = FALSE)

B <- 2

y_1 <- ((-10*100):0)/100
y_2 <- 2*(-y_1)^(2/3)

polygon(x = c(y_1, 0, -10, -10), y=c(y_2, -10, -10, 10), col="red", border = "red")

lines(x=y_1, y=y_2)

abline(h=0, v=0)

abline(v=-1, h=B, lty=2)
text("y_2 = B", x=5, y=3)
text("y_1 = -1", x=-2.5, y=7)

box()

```

(b) Solve the firm's profit maximization problem to find $\pi(p)$ and $Y^*(p)$. (It may help to set $z=-y_1$ as the amount of input used, explain why a profit-maximizing firm will set $y_2 = Bz^{\frac{2}{3}}$, and solve a single-dimensional maximization problem for $z$, but be sure to state your solution $Y^*(p) \in R^2$.)

(c) Since $Y*(p)$ is single-valued, I'll refer to it below as $y(p)$.  Verify that $\pi(\cdot)$ is homogeneous of degree 1, and $y(\cdot)$ is homogeneous of degree 0.

(d) Verify that $y_1(p) = \frac{\partial\pi}{\partial p_1}(p)$ and $y_2(p) = \frac{\partial\pi}{\partial p_2}(p)$.

(e) Calculate $D_p y(p)$, and verify it is symmetric, positive semidefinite, and $[D_p y]p = 0$

\pagebreak

# Question 3 ...and recovering technology from the profit function

Finally, suppose we didn't know a firm's production set $Y$, but did know its profit function was

$$
\pi(p) Ap_1^{-2}p_2^3
$$
for all $p_1, p_2 > 0$ and $A > 0$ a known constant.

(a) What conditions must hold for this profit function to be rationalizable?  (You don't need to check them.)

(b) Recall that the outer bound was defined

