---
title: "ECON 711 - PS 5"
author: "Alex von Hafften^[I worked on this problem set with a study group of Michael Nattinger, Andrew Smith, Tyler Welch, and Ryan Mather. I also discussed problems with Emily Case, Sarah Bass, and Danny Edgel.]"
date: "10/12/2020"
output: pdf_document
header-includes:
- \newcommand{\N}{\mathbb{N}}
- \newcommand{\Z}{\mathbb{Z}}
- \newcommand{\R}{\mathbb{R}}
- \newcommand{\Q}{\mathbb{Q}}
- \usepackage{bm}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Question 1. The Consumer Problem

Solve the Consumer Problem and state the Marshallian demand $x(p,w)$ and indirect utility $v(p, w)$ for the following utility functions:\footnote{For parts (e) and (f), you may describe the Marshallian demand in words rather than giving mathematical formulas if you prefer, and you can ignore the "knife-edge" cases where two prices or sums of prices are exactly equal, but you should still give formulas for the indirect utility function.}

(a) $u(x) = x_1^\alpha + x_2^\alpha$ for $\alpha<1$\footnote{This answer assumes $\alpha>0$.}

The consumer problem is $\max\{x_1^\alpha + x_2^\alpha\}$ subject to $p_1x_1+p_2x_2 \le w$, $x_1 \ge 0$, and $x_2 \ge 0$. Based on the utility function, it is clear that $x_1^* = x_2^* = 0$ is impossible because the consumer's utility is zero and they would be infinitely better off if they consume $\varepsilon$ of either $x_1$ or $x_2$. Thus, let us assume $x_1^*>0$.\footnote{This assumption is for expositional clarity.  Since the utility function is symmetric, results do not change if we had assumed $x_2^*>0$.} Notice that $u$ is differentiable and concave in $x_i$: $\frac{\partial^2 u}{\partial^2 x_i} = \alpha (\alpha-1) x_i^{\alpha-2} < 0$. So, by Theorem 2 in lecture 10 notes, if $(x^*, \lambda^*, \mu^*)$ satisfies the Kuhn-Tucker conditions, $x^*$ solves the consumer problem.  The Legrangian is $\mathcal{L}(x, \lambda, \mu) = (x_1^\alpha + x_2^\alpha) + \lambda(w - p_1x_1 - p_2x_2) + \mu_1x_1 + \mu_2x_2$. The Kuhn-Tucker FOC are

$$
\frac{\partial \mathcal{L}}{\partial x_i} =  0 
\implies \alpha (x_i^*)^{\alpha-1} - \lambda^* p_i + \mu^*_i = 0 
\implies x_i^* = \Bigg(\frac{\lambda^* p_i - \mu_i^*}{\alpha}\Bigg)^\frac{1}{\alpha-1}
$$

Since $x_1^*>0$, the above equation implies that $\lambda^*>0$. By complementary slackness, $\lambda^*>0 \implies w=p_1x_1^*+p_2x_2^* \implies x^*_2 = \frac{p_1}{p_2}x_1^* > 0$. Thus, again by complementary slackness, $x_1^*> 0$ and $x_2^*> 0 \implies \mu_1^*=\mu_2^*=0$.

$$
\Bigg(\frac{\lambda^* p_1}{\alpha}\Bigg)^\frac{1}{\alpha-1} + \frac{p_1}{p_2}\Bigg(\frac{\lambda^* p_1}{\alpha}\Bigg)^\frac{1}{\alpha-1} = w \implies
\lambda^* = \frac{\alpha}{p_1}\Bigg(\frac{p_2w}{p_1+p_2}\Bigg)^{(\alpha-1)}
$$

$$
x^*_1 
= \Bigg(\frac{\frac{\alpha}{p_1}\Big(\frac{p_2w}{p_1+p_2}\Big)^{(\alpha-1)} p_1}{\alpha}\Bigg)^\frac{1}{\alpha-1}
=\frac{p_2}{p_1+p_2}w 
\implies 
x_2^* = \frac{p_1}{p_2}\frac{p_2}{p_1+p_2}w
=\frac{p_1}{p_1+p_2}w
$$

\pagebreak

Thus, the Marshallian demand and the indirect utility are

\begin{align*}
x(p, w) &= \Bigg( \frac{p_2}{p_1+p_2}w , \frac{p_1}{p_1+p_2}w \Bigg) \\
v(p, w) &= \Bigg( \frac{p_2}{p_1+p_2}w\Bigg)^\alpha + \Bigg( \frac{p_1}{p_1+p_2}w\Bigg)^\alpha 
= \frac{(p_1^\alpha+p_2^\alpha)w^\alpha}{(p_1+p_2)^\alpha}
\end{align*}

(b) $u(x) = x_1+x_2$

The consumer problem is $\max\{x_1 + x_2\}$ subject to $p_1x_1+p_2x_2 \le w$, $x_1 \ge 0$, and $x_2 \ge 0$. Notice that although $u$ is differentiable, it is not concave in $x_i$: $\frac{\partial^2 u}{\partial^2 x_i} = 0$. Thus, $(x^*, \lambda^*, \mu^*)$ satisfying the Kuhn-Tucker conditions do not guarantee a solution to the consumer problem.  The linear utility function can be represented as straight indifference lines with slopes of -1 (see figure for examples with blue indifference curves, green budget constraints, and red solutions). Since the budget constraint is also a straight line, the consumer chooses corner solutions when $p_1 \neq p_2$.  When $p_1 < p_2$, the consumer can afford more of $x_1$, so she buys $x_1 = \frac{w}{p_1}$ and none of $x_2$. When $p_1 > p_2$, the consumer can afford more of $x_2$, so she buys $x_2 = \frac{w}{p_2}$ and none of $x_1$.  When $p_1 = p_2$, there is a continuum of solutions along the overlaid indifference curve and budget constraint.

The Marshallian demand and the indirect utility are

\begin{align*}
x(p, w) 
= \begin{cases} (w/p_1, 0) & \text{ if } p_1 < p_2 \\
                (0, w/p_2) & \text{ if } p_1 > p_2 \\
                \{(tw/p_1, (t-1)w/p_1) \; \forall t \in[0, 1]\} & \text{ if } p_1 = p_2
\end{cases}
\end{align*}

\begin{align*}
v(p, w) 
= \begin{cases} w/p_1 & \text{ if } p_1 \le p_2 \\
                w/p_2 & \text{ if } p_1 > p_2 
  \end{cases}
\end{align*}

``` {r, echo = FALSE, fig.height = 2.75, fig.width = 6.5}
par(mfrow = c(1, 3))

plot(1, type = "n", xlim=c(0, 10), ylim =c(0, 10), xaxs = "i", yaxs = "i", xlab = "x_1", ylab = "x_2", main = "p_1 < p_2")

for (i in 1:20) abline(a = i, b = -1, col = "blue")

abline(a = 4, b = -1/2, lwd = 2, col = "green")
points(x=8, y=0, pch = 19, col="red")

plot(1, type = "n", xlim=c(0, 10), ylim =c(0, 10), xaxs = "i", yaxs = "i", xlab = "x_1", ylab = "x_2", main = "p_1 > p_2")

for (i in 1:20) abline(a = i, b = -1, col = "blue")

abline(a = 8, b = -2, lwd = 2, col = "green")
points(x=0, y=8, pch = 19, col="red")

plot(1, type = "n", xlim=c(0, 10), ylim =c(0, 10), xaxs = "i", yaxs = "i", xlab = "x_1", ylab = "x_2", main = "p_1 = p_2")

for (i in 1:20) abline(a = i, b = -1, col = "blue")

abline(a = 8, b = -1, lwd = 4, col = "red")
abline(a = 8, b = -1, col = "green")
```

\pagebreak

(c) $u(x) = x_1^\alpha + x_2^\alpha$ for $\alpha>1$

The consumer problem is $\max\{x_1^\alpha + x_2^\alpha\}$ subject to $p_1x_1+p_2x_2 \le w$, $x_1 \ge 0$, and $x_2 \ge 0$. Notice that although $u$ is differentiable, it is not concave in $x_i$: $\frac{\partial^2 u}{\partial^2 x_i} = \alpha (\alpha-1) x_i^{\alpha-2} \ge 0$. Thus, $(x^*, \lambda^*, \mu^*)$ satisfying the Kuhn-Tucker conditions do not guarantee a solution to the consumer problem.  As shown in the figure, the indifference curves when $\alpha > 1$ become concave to the origin: $x_i = f(x_j) = (\bar{u} - x_j^\alpha)^{1/\alpha} \implies \frac{\partial^2 f}{\partial^2x_i}$

``` {r, echo = FALSE, fig.height = 2.75, fig.width = 6.5}
par(mfrow = c(1, 3))

x <- (0:(10*100))/100

plot(1, type = "n", xlim=c(0, 10), ylim =c(0, 10), xaxs = "i", yaxs = "i", xlab = "x_1", ylab = "x_2", main = "alpha = 1.5")

for (u_bar in (1:30)*2) lines(x=x, y=(u_bar - x^1.5)^(1/1.5), col = "blue")

plot(1, type = "n", xlim=c(0, 10), ylim =c(0, 10), xaxs = "i", yaxs = "i", xlab = "x_1", ylab = "x_2", main = "alpha = 5")

x <- (0:(10*10000))/10000

for (u_bar in exp(1:30)) lines(x=x, y=(u_bar - x^5)^(1/5), col = "blue")

plot(1, type = "n", xlim=c(0, 10), ylim =c(0, 10), xaxs = "i", yaxs = "i", xlab = "x_1", ylab = "x_2", main = "alpha = infinity")

for (i in 1:20) {
  segments(x0 = 0, y0 = i,
           x1 = i, y1 = i, 
           col = "blue")
    segments(x0 = i, y0 = 0,
           x1 = i, y1 = i, 
           col = "blue")
}
```

(d) $u(x) = \min\{x_1, x_2\}$ (Leontief utility)



(e) $u(x) = \min\{x_1 + x_2, x_3 + x_4\}$



(f) $u(x) = \min\{x_1, x_2\} + \min\{x_3, x_4\}$



\pagebreak

# Question 2. CES Utility

Throughout this problem, let $X=\R^k_+$, and let $(a_1, a_2, ..., a_k)$ be a set of strictly positive coefficients which sum to 1. You may assume prices and wealth are strictly positive, and ignore cases where two or more prices are identical.

(a) For each of the following utility functions, solve the consumer problem and state $x(p, w)$:

i. linear utility $u(x)=x_1+x_2+...+x_k$
ii. Cobb-Douglas utility $u(x)=x_1^{a_1}x_2^{a_2}...x_k^{a_k}$
iii. Leontief utility $u(x)=\min\{\frac{x_1}{a_1}, \frac{x_2}{a_2}, ..., \frac{x_k}{a_k}\}$

(b) Consider the Constant Elasticity of Substitution (CES) utility function $u(x) = \Bigg( \sum_{i=1}^k a_i^{\frac{1}{s}} x_i^\frac{s-1}{s} \Bigg)^\frac{s}{s-1}$ with $s \in (0,1) \cup(1, +\infty)$. Solve the consumer problem and state $x(p,w)$.\footnote{Recall that maximizing a function $(f(x))^\frac{s}{s-1}$ is the same as maximizing $f(x)$ when $s > 1$, and the same as minimizing $f(x)$ when $s < 1$.}

(c) Show that CES utility gives the same demand as linear utility in the limit $s \to + \infty$, as Cobb-Douglas utility in the limit $s \to 1$, and as Leontief utility in the limit $s \to 0$.

(d) The Elasticity of Substitution between goods 1 and 2 is defined as $\xi_{1,2} = - \frac{\partial\log\Big(\frac{x_1(p,w)}{x_2(p,w)}\Big)}{\partial log\Big(\frac{p_1}{p_2}\Big)}=- \frac{\partial\Big(\frac{x_1(p,w)}{x_2(p,w)}\Big)}{\partial \Big(\frac{p_1}{p_2}\Big)}\frac{\frac{p_1}{p_2}}{\frac{x_1(p,w)}{x_2(p,w)}}$ While this looks complicated, in the case of CES demand, we can write the ratio $\frac{x_1}{x_2}$ as a relatively simple function of the price ratio $\frac{p_1}{p_2}$, and calculate this elasticity without much difficulty. Calculate the elasticity of substitution for CES demand, note its value as $s \to + \infty$, $s \to 1$, and $s \to 0$.


\pagebreak

# Question 3. Exchange Economies

We've been considering the problem facing a consumer with wealth $w$ at prices $p$. An "exchange economy" is different model where instead of money, each consumer is endowed with an initial bundle of goods $e \in \R^k_+$, and can either buy or sell any quantity of the goods at market prices $p$. The consumer's problem is then $\max_{x \in \R_+^k} u(x)$ subject to $p \cdot x \ge p \cdot e$ (i.e., the consumer's "budget" is the market value of the goods they start with).  Assume preferences are locally non-satiated and the consumer's problem has a unique solution $x(p, e)$. We'll say the consumer is a net buyer of good $i$ if $x_i(p, e) > e_i$ and a net seller if $x_i(p, e) < e_i$.

(a) Show that if $p_i$ increases, the consumer cannot switch from being a net seller to a net buyer.



(b) Suppose $u$ is differentiable and concave. Use the Legrangian and the envelope theorem to show that $\frac{\partial v}{\partial p_i}$ is negative if the consumer is a net buyer of good $i$, and positive if the consumer is a net seller.



(c) Consider the following statement. "If the consumer is a net buyer of good $i$ and its price goes up, the consumer must be worse off." True or false? Explain.

