---
title: "ECON 713B - Problem Set 1"
author: Alex von Hafften^[I worked on this problem set with a study group of Michael
  Nattinger, Andrew Smith, and Ryan Mather. I also discussed problems with Sarah Bass,
  Emily Case, Danny Edgel, and Katherine Kwok.]
date: "4/1/2021"
output:
  pdf_document: default
  html_document:
    df_print: paged
header-includes:
- \newcommand{\N}{\mathbb{N}}
- \newcommand{\Z}{\mathbb{Z}}
- \newcommand{\R}{\mathbb{R}}
- \newcommand{\Q}{\mathbb{Q}}
- \newcommand{\Lf}{\mathcal{L}}
- \usepackage{bm}
- \usepackage{bbm}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
```

## 1 All Pay Auction

Consider a symmetric IPV (independent private values) setting with $N$ bidders. Find an equilibrium of the all-pay auction when each bidder's valuation is an iid draw from $F(x) = x^a$ for $a \in (0, \infty)$ and $x \in [0, 1]$.

(a) Define this auction as a Bayesian game.

A Bayesian game is a five-tuple $(I, \{S_i\}, \{u_i(\cdot)\}, \Theta, F(\cdot))$:

- The set of players is $I = \{1, ..., N\}$.
- The action set of player $i \in I$ is $S_i = [0, \infty)$.
- The payoff for player $i \in I$ is 

$$
u_i(b_1, ..., b_N;v_1, ..., v_N) 
= u_i(b_1, ..., b_N;v_i)
=
\begin{cases}
v_i - b_i & \text{if } b_i > \max_{j \neq i} b_{j} \\
\frac{1}{2}(v_i - b_i) + \frac{1}{2}(-b_i) & \text{if } b_i = \max_{j \neq i} b_{j} \\
-b_i & \text{if } b_i < \max_{j \neq i} b_{j} \\
\end{cases}
$$

- $\Theta = [0, 1] \times ... \times [0, 1]$.
- $F(x) = x^a$ for $a \in (0, \infty)$

(b) Find equilibrium strategies of all players.

Focus on BNE with symmetric, strictly increasing, and differentiable bids $b(v_i)$. Since the $F$ is continuous and $b(v_i)$ is strictly increasing, the probability of a tie is zero.  The expected payoff for bidder $i$ is:

\begin{align*}
E[u_i(b_1, ..., b_N; v_i)]
&= (v_i - b_i) \Pr(b_i > b_j, \forall j\neq i) + (-b_i)\Pr(b_i < b_j, \forall j\neq i)\\
&= v_i \Pr(b_i > b_j, \forall j\neq i) - b_i
\end{align*}

Suppose bidder $j \neq i$ submit $b(v_j)$:

\begin{align*}
\Pr(b_i > b_j, \forall j\neq i)
&= \Pr(b(v_i) > b(v_j), \forall j\neq i)\\
&= \Pr(b^{-1}(b(v_i)) > v_j, \forall j\neq i)\\
&=^{iid} F(b^{-1}(b(v_i)))^{N-1}\\
&= ((b^{-1}(b(v_i)))^a)^{N-1}\\
&= (b^{-1}(b(v_i)))^{aN-a}
\end{align*}

Thus, the expected payoff of bidder $i$ is:

$$
E[u_i(b_1, ..., b_N; v_i)] = v_i (b^{-1}(b(v_i)))^{aN-a} - b(v_i)
$$

FOC [$b(v_i)$]:

\begin{align*}
0 &= (aN-a)v_i (b^{-1}(b(v_i)))^{aN-a-1}\frac{1}{b'(v_i)} - 1 \\
\implies b'(v_i) &= (aN-a)v_i^{aN-a}\\
\implies b(v_i) &= \frac{aN-a}{aN-a+1}v_i^{aN-a+1} + c_i
\end{align*}

For $v_i = 0$, bidder $i$ would bid zero: $b(v_i) = 0 \implies c_i = 0$.

$$
\implies b(v_i) = \frac{aN-a}{aN-a+1}v_i^{aN-a+1}
$$

Since $aN - a + 1 > 0$ and $\frac{aN-a}{aN-a+1}$, $b$ is strictly increasing.

(c) Verify that the strategies that you have found do constitute an equilibrium.

We can verify that $b$ is an equilibrium strategy by verifying that $b(v_i)$ is the best response for player $i$ when bidders $j \neq i$ bid $b(v_j)$. 

\begin{align*}
E[u_i(b_1, ..., b_N; v_i)]
&= v_i \Pr\Bigg(b_i > \frac{aN-a}{aN-a+1}v_j^{aN-a+1}, \forall j\neq i\Bigg) - b_i\\
&= v_i \Pr\Bigg(\Bigg(\frac{aN-a+1}{aN-a}b_i \Bigg)^{\frac{1}{aN-a+1}} > v_j, \forall j\neq i\Bigg) - b_i \\
&= v_i \Bigg(\frac{aN-a+1}{aN-a}b_i \Bigg)^{\frac{aN-a}{aN-a+1}} - b_i
\end{align*}

FOC [$b_i$]:

\begin{align*}
0 &= v_i \frac{aN-a}{aN-a+1}\Bigg(\frac{aN-a+1}{aN-a}b_i \Bigg)^{\frac{aN-a}{aN-a+1}-1}\frac{aN-a+1}{aN-a} - 1\\
\implies
\frac{1}{v_i} &= \Bigg(\frac{aN-a+1}{aN-a}b_i \Bigg)^{\frac{-1}{aN-a+1}}\\
\implies 
b_i &= b(v_i) = \frac{aN-a}{aN-a+1} v_i^{aN-a+1}
\end{align*}

Thus, all bidders playing $b(\cdot)$ is an equilibrium.

\pagebreak

(d) Does the bidding become more competitive when $a$ increases? Explain.

Conditional on $v_i$ and $N$, a higher $a$ results in more mass closer to 1.  Thus, the probability that bidder $i$ wins the auction decreases, so bidder $i$ should decrease her bid.  This makes the bidding less competitive.  Unconditional on $v_i$, a higher $a$ results in higher realizations of $v_i$, so the bids are correspondingly larger.

Consider the derivative of the bidder function with respect to $a$

\begin{align*}
\frac{\partial}{\partial a}b(v_i) 
&= \frac{\partial}{\partial a}\Bigg(\frac{aN-a}{aN-a+1}v_i^{aN-a+1}\Bigg)\\
&= \frac{N-1}{aN-a+1}v_i^{aN-a+1} 
- \frac{(aN-a)(N-1)}{(aN-a+1)^2}v_i^{aN-a+1} 
+ \frac{(N-1)(aN-a)}{aN-a+1}v_i^{aN-a+1}\log(v_i)
\end{align*}

The derivative is not strictly positive or negative, so increasing $a$ may increase competition and may decrease competition. But for a sufficiently large $a$, the derivative is negative, so an increase in $a$ decreases a bid conditional on $v_i$.

We can see that a higher $a$ reduces bids for a sufficiently large $a$ in the figure below ($N=5, a \in (0, 2), v_i \in \{0.25, 0.5, 0.75\}$):

```{r problem_1d, echo = FALSE}

N <- 5

v <- c(.25, .5, .75)

a <- seq(.001, 2, .001)

plot(1, xlim = c(0, 2), ylim = c(0, .4), type = "n", xlab = "a", ylab = "b(v_i)", main = "N = 5")

lines(x = a, y = (a*N-a) / (a*N-a+1) * v[1]^(a*N-a+1), col = "black")
lines(x = a, y = (a*N-a) / (a*N-a+1) * v[2]^(a*N-a+1), col = "blue")
lines(x = a, y = (a*N-a) / (a*N-a+1) * v[3]^(a*N-a+1), col = "red")

text("v_i = 0.25", col = "black", x= .25, y=0)
text("v_i = 0.5", col = "blue", x= .5, y=.15)
text("v_i = 0.75", col = "red", x= 1, y=.25)

```

\pagebreak

(e) Compute the expected payment from each bidder before and after she learns her value.

The expected payment from bidder $i$ conditional on $v_i$ is their bid:

$$
b(v_i) = \frac{aN-a}{aN-a+1}v_i^{aN-a+1}
$$

The expected payment from bidder $i$ unconditionally is:

\begin{align*}
b(v_i) 
&= \int_0^1 \frac{aN-a}{aN-a+1}v_i^{aN-a+1} a v_i^{a-1} dv_i\\
&= a \frac{aN-a}{aN-a+1} \int_0^1 v_i^{aN} dv_i\\
&= a \frac{aN-a}{aN-a+1} \Bigg[\frac{1}{aN+1} v_i^{aN+1}\Bigg]_0^1  \\
&= \frac{a^2N-a^2}{(aN-a+1)(aN+1)}
\end{align*}

The expected payoff of bidder $i$ conditional on $v_i$ is:

\begin{align*}
E[u_i(v_i)] 
&= v_i^{aN-a+1} - b(v_i) \\
&= v_i^{aN-a+1} - \frac{aN-a}{aN-a+1}v_i^{aN-a+1}\\
&= \frac{v_i^{aN-a+1}}{aN-a+1}
\end{align*}

The unconditional expected payoff of bidder $i$ is:

\begin{align*}
\int_0^1 E[u_i(v_i)] f(v_i) dv_i 
&= \int_0^1 \frac{v_i^{aN-a+1}}{aN-a+1} av_t^{a-1} dv_i \\
&= \int_0^1 \frac{a v_i^{aN}}{aN-a+1} dv_i \\
&= \Bigg[ \frac{a v_i^{aN+1}}{(aN-a+1)(aN+1)} \Bigg]_0^1 \\
&= \frac{a}{(aN-a+1)(aN+1)} 
\end{align*}

\pagebreak

## 2 Tricky Seller

Two people are interested in one object. Their valuations are drawn independently from $F(x) = x$ and $F(x) = x^2$, respectively, with $x \in [0, 1]$. The seller's value (a cost, perhaps) for the object is known, $c \in [0, 1]$.

(a) Describe outcome of the First-Price Auction with a reserve price $r$.

Let us disregard the reserve price for the moment and consider the auction as Bayesian game $(I, \{S_i\}, \{u_i(\cdot)\}, \Theta, F(\cdot))$:

- The set of players is $I = \{1, 2\}$.
- The action set of player $i \in I$ is $B_i = [0, \infty)$.
- The payoff for player $i \in I$ is 

$$
u_i(b_1, b_2; x_1, x_2) 
= u_i(b_1, b_2; x_i)
=
\begin{cases}
x_i - b_i              & \text{if } b_i > b_{j} \\
\frac{1}{2}(x_i - b_i) & \text{if } b_i = b_{j} \\
0                      & \text{if } b_i < b_{j} \\
\end{cases}
$$

- $\Theta = [0, 1] \times [0, 1]$.
- $F_1(x) = x$ and $F_2(x) = x^2$.

The expected payoff of bidder 1 when bidder 2 plays $b_2(x_2) = a_2x_2$ (linear bid function and bids zero when value is zero):

\begin{align*}
E[u_1(b_1, b_2; x_1)]
&= (x_1 - b_1) \Pr(b_1 > b_2(x_2))\\
&= (x_1 - b_1) \Pr(b_2^{-1}(b_1) > x_2)\\
&= (x_1 - b_1) F_2(b_2^{-1}(b_1))\\
&= (x_1 - b_1) (b_2^{-1}(b_1))^2\\
&= (x_1 - b_1) (\frac{b_1}{a_2})^2 \\
&= \frac{x_1 b_1^2}{a_2^2}- \frac{b_1^3}{a_2^2}
\end{align*}

FOC:

$$
2\frac{x_1 b_1}{a_2^2}- 3\frac{b_1^2}{a_2^2} = 0 \\
\implies
b_1(x_1) = \frac{2}{3}x_1
$$

The expected payoff of bidder 2 when bidder 1 plays $b_1(x_1) = a_1x_1$ (linear bid function and bids zero when value is zero):

\begin{align*}
E[u_2(b_1, b_2; x_2)]
&= (x_2 - b_2) \Pr(b_2 > b_1(x_1))\\
&= (x_2 - b_2) \Pr(b_1^{-1}(b_2) > x_1)\\
&= (x_2 - b_2) F_1(b_1^{-1}(b_2))\\
&= (x_2 - b_2) b_1^{-1}(b_2)) \\
&= (x_2 - b_2) \frac{b_2}{a_2} \\
&= \frac{b_2x_2}{a_2} - \frac{b_2^2}{a_2}
\end{align*}

$$
$$

FOC:

$$
\frac{x_2}{a_2} - 2\frac{b_2}{a_2} = 0 \\
\implies
b_2(x_2) = \frac{x_2}{2}
$$

Thus, bidder 1 bids $b_1(x_1) = \frac{2}{3}x_1$ and $b_2(x_2) = \frac{x_2}{2}$, so both bidders underbid their values.  

The seller will only sell the good if the revenue weakly exceeds $c$, so the reservation price $r=c$ and it is common knowledge.  The bidders know this, so $r$ is a floor on the amount that they will bid:

$$
b_1(x) = 
\begin{cases}
\frac{2}{3}x,   & \frac{2}{3}x \ge r \\
r,              & x_1 \ge r > \frac{2}{3}x \\
0,              & r >x
\end{cases}
$$

$$
b_2(x) = 
\begin{cases}
\frac{1}{2}x,   &  \frac{1}{2}x \ge r \\
r,              & x \ge r > \frac{1}{2}x \\
0,              & r >x
\end{cases}
$$

- If both bids are below $r$, then the both bidders submit zero bids and the seller doesn't sell.
- If one or both bids is above $r$, then a bid at least as large as $r$ is submitted.
- Bidders still would have positive surplus if $r$ is between their value and bid absent a reserve price, so they would bid the reserve price.

(b) Describe outcome of the Second-Price Auction with a reserve price $r$.

As discussed in lecture, bidding $b(v)=v$ is a weakly dominate strategy. Similar to FPA, the seller will only sell the good if the revenue weakly exceeds $c$, so the reservation price $r=c$ and it is common knowledge.  The bidders know this, so $r$ is a floor on the amount that they will bid:

$$
b_1(x) = 
b_2(x) = 
\begin{cases}
x,    & x \ge r \\
0,    & x < r
\end{cases}
$$

- If both bids or one bid are below $r$, then the auction revenue is below the reserve price and the seller doesn't sell.
- If both bids is above $r$, then a bid at least as large as $r$ is submitted.

(c) What auction and what $r$ will the seller choose? Which player wins more often?

As mentioned above, the seller will choose $r = c$.  If $r > c$, the seller may turn down trades that they would be better off if they accepted.  If $r < c$, they may be selling the good for less than they value it.

The seller will choose the auction format that maximizes their revenue...

Bidder 2 is more likely to have higher valuation, so they are more likely to submit a higher bid.

(d) Suppose now that $c = 0$ and there is no reserve price. Suppose that a seller can offer discount of $\alpha$ to one of the bidders in the second-price auction. If a bidder is offered a discount $\alpha \in [0; 1]$, then, if she wins, she pays only a fraction $\alpha$ of what she had to pay otherwise. Who should be offered a discount? Compute the optimal discount and expected revenues.

...

\pagebreak

## 3 Third Price Auction

Consider a third-price auction with three players: an auction in which bidder with the highest value wins, but pays only the third highest bid. Assume that valuation of players are iid from the uniform distribution on [0, 1].

(a) Define the auction as a Bayesian game.

For this part, I consider a third-price auction with only three players. A Bayesian game is a five-tuple $(I, \{S_i\}, \{u_i(\cdot)\}, \Theta, F(\cdot))$:

- The set of players is $I = \{1, 2, 3\}$.
- The action set of player $i \in I$ is $B_i = [0, \infty)$.
- The payoff for player $i \in I$ is 

$$
u_i(b_1, b_2, b_3; v_1, v_2, v_3) 
= u_i(b_1, b_2, b_3; v_i)
=
\begin{cases}
v_i - b_k              & \text{if } b_i > b_j \ge b_k, \\
\frac{1}{3}(v_i - b_k) & \text{if } b_i = b_j = b_k, \\
\frac{1}{2}(v_i - b_k) & \text{if } b_i = b_j > b_k, \\
0                      & \text{otherwise.}
\end{cases}
$$

- $\Theta = [0, 1] \times [0, 1] \times [0, 1]$.
- $F(v) = v$.

(b) Prove that a bid of $b_i(v_i) = \frac{n - 1}{n - 2} v_i$ is a symmetric Bayes Nash equilibrium of the third-price auction.

In this part, I consider a third-price auction with $n$ bidders. I show that bidder $i$'s best response to $b(v_{-i})  = \frac{n - 1}{n - 2} v_{-i}$ is to play $b(v_i) = \frac{n - 1}{n - 2} v_i$ below and thus it is a symmetric Bayes Nash equilibrium.  The expected payoff of bidder $i$ is

\begin{align*}
E[u_i(b_1, ..., b_n; v_i)] 
&= (v_i - E[b_{(n-2)}| b_i > b_j, j \neq i]) \Pr(b_i > b_j, j \neq i)\\
&= \Bigg(v_i - \frac{n - 1}{n - 2} E\Bigg[v_{(n-2)}| b_i > \frac{n - 1}{n - 2} v_j, j \neq i\Bigg]\Bigg) \Pr\Bigg(b_i > \frac{n - 1}{n - 2} v_j, j \neq i\Bigg)\\
&= \Bigg(v_i - \frac{n - 1}{n - 2} E\Bigg[v_{(n-2)} \Bigg| \frac{n - 2}{n - 1} b_i > v_j, j \neq i\Bigg]\Bigg) \Pr\Bigg(\frac{n - 2}{n - 1} b_i > v_j, j \neq i\Bigg)\\
&= \Bigg(v_i - \frac{n - 1}{n - 2} E[w_{(n-2)} ]\Bigg) F\Bigg(\frac{n - 2}{n - 1} b_i\Bigg)^{n-1}\\
&= \Bigg(v_i - \frac{n - 1}{n - 2} \frac{n - 2}{n - 1} b_i\frac{n - 2}{n} \Bigg) \Bigg(\frac{n - 2}{n - 1} b_i\Bigg)^{n-1}\\
&= \Bigg(v_i - \frac{n - 2}{n} b_i\Bigg) \Bigg(\frac{n - 2}{n - 1} b_i\Bigg)^{n-1}\\
&= \Bigg(\frac{n - 2}{n - 1}\Bigg)^{n-1} v_i b_i^{n-1} - \frac{n - 2}{n}\Bigg(\frac{n - 2}{n - 1}\Bigg)^{n-1}b_i^n
\end{align*}

where $w_j \sim U(0, \frac{n - 2}{n - 1} b_i)$ for $j \neq i$.  Generally, note that if $X_1, ..., X_n \sim U(0, 1)$, then the $k$th order statistic $X_{(k)} \sim Beta(k, n - k +1) \implies E[X_{(k)}] = \frac{k}{n+1}$. So, $E[w_{(n-2)} ] = \frac{n - 2}{n - 1} b_i\frac{n - 2}{n}$.

\pagebreak

FOC [$b_i$]:

$$
(n-1)\Bigg(\frac{n - 2}{n - 1}\Bigg)^{n-1} v_i b_i^{n-2} = n\frac{n - 2}{n}\Bigg(\frac{n - 2}{n - 1}\Bigg)^{n-1}b_i^{n-1} \\
\implies
b_i(v_i)  = \frac{n-1}{n - 2}v_i
$$

Thus, $b_i(v_i)  = \frac{n-1}{n - 2}v_i$ is a best response.

(c) Show that the expected revenue of a seller in the third-price auction is $R_3 = \frac{n - 1}{n + 1}$.

The expected seller revenue is the expected value of the third highest bid:

\begin{align*}
R_3 
&= E[b(v_{(n-2)})] \\
&= E\Bigg[\frac{n - 1}{n - 2} v_{(n-2)}\Bigg] \\
&= \frac{n - 1}{n - 2} E[ v_{(n-2)} ] \\
&= \frac{n - 1}{n - 2} \frac{n - 2}{n + 1} \\
&= \frac{n - 1}{n + 1} 
\end{align*}

(d) What is the symmetric Bayes-Nash equilibrium strategy in a $k$th price auction? (You need only state how each bidder bids; you need not provide a detailed analysis.)

From lecture notes and this problem, we know the bidding function in symmetric BNEs for $k \in \{1, 2, 3\}$:

$$
b (v_i) = 
\begin{cases}
\frac{n-1}{n} v_i   & k = 1\\
v_i                 & k = 2\\
\frac{n-1}{n-2} v_i & k = 3
\end{cases}
$$

These findings suggest that $b(v_i) =\frac{n-1}{n-k+1}$ for all $k \in \N$ is a candidate.